---
title: "Estimate Cor heter"
author: "Yuxin Zou"
date: 2018-12-3
output: 
  workflowr::wflow_html:
    code_folding: hide
---

```{r}
rrmse = function(data, model){
  sqrt(mean((data$B - model$result$PosteriorMean)^2)/mean((data$B - data$Bhat)^2))
}

ROC.table = function(data, model){
  sign.test = data*model$result$PosteriorMean
  thresh.seq = seq(0, 1, by=0.005)[-1]
  m.seq = matrix(0,length(thresh.seq), 2)
  colnames(m.seq) = c('TPR', 'FPR')
  for(t in 1:length(thresh.seq)){
    m.seq[t,] = c(sum(sign.test>0 & model$result$lfsr <= thresh.seq[t])/sum(data!=0),
                  sum(data==0 & model$result$lfsr <=thresh.seq[t])/sum(data==0))
  }
  return(m.seq)
}

library(knitr)
library(kableExtra)
```

## Common noise correlation

We simulate null data which has common noise correlation structure. We fit mash model without and with the estimated correlation structure. There are lots of false positives in the model without the correlation structure. The posterior mean is far from the truth.

```{r}
library(mvtnorm)
library(mashr)
set.seed(1)
n = 10000; p = 5
B = matrix(0,n,p)
V = matrix(0.75, p, p); diag(V) = 1
Bhat = rmvnorm(n, sigma = V)
simdata = list(B = B, Bhat = Bhat, Shat = 1)
```

```{r}
data = mash_set_data(Bhat, Shat=1)
U.c = cov_canonical(data)
m.ignore = mash(data, U.c, verbose = FALSE)

V.mV = estimate_null_correlation(data, U.c, details = T)
m.mV = V.mV$mash.model

data.true = mash_update_data(data, V = V)
m.true = mash(data.true, U.c, verbose = FALSE)
```

```{r}
ign = c(get_loglik(m.ignore), length(get_significant_results(m.ignore)))

mV = c(get_loglik(m.mV), length(get_significant_results(m.mV)))

true = c(get_loglik(m.true), length(get_significant_results(m.true)))

tmp = rbind(ign, mV, true)
row.names(tmp) = c('ignore', 'mV', 'true')
colnames(tmp) = c('loglik', '# signif')
tmp %>% kable() %>% kable_styling()
```

RRMSE:
```{r}
tmp = c(rrmse(simdata, m.ignore), rrmse(simdata, m.mV), rrmse(simdata, m.true))
barplot(tmp, names.arg = c('ignore', 'mV', 'true'))
```

## Two different noise correlations

Now, we simulate data with 2 noise correlation structures. Half of the null data have no noise correlation, the other half have noise correlation.
```{r}
Bhat1 = rmvnorm(n/2, sigma = diag(p))
Bhat2 = rmvnorm(n/2, sigma = V)
Bhat = rbind(Bhat1, Bhat2)
simdata = list(B = B, Bhat = Bhat, Shat = 1)

data = mash_set_data(Bhat, Shat=1)
U.c = cov_canonical(data)
m.I = mash(data, U.c, verbose = FALSE)

data.V = mash_update_data(data, V = V)
m.V = mash(data.V, U.c, verbose = FALSE)

V.mV = estimate_null_correlation(data, U.c, details = T)
m.mV = V.mV$mash.model

Vtrue = array(0,dim=c(p,p,n))
Vtrue[,,1:(n/2)] = diag(p)
Vtrue[,,(n/2+1): n] = V
data.true = mash_update_data(data, V = Vtrue)
m.true = mash(data.true, U.c, verbose = FALSE, algorithm.version = 'R')
```

The estimated V is
```{r}
V.mV$V
```

```{r}
Ionly = c(get_loglik(m.I), length(get_significant_results(m.I)))

Vonly = c(get_loglik(m.V), length(get_significant_results(m.V)))

mV = c(get_loglik(m.mV), length(get_significant_results(m.mV)))

true = c(get_loglik(m.true), length(get_significant_results(m.true)))

tmp = rbind(Ionly, Vonly, mV, true)
row.names(tmp) = c('I only', 'V only', 'mV', 'true')
colnames(tmp) = c('loglik', '# signif')
tmp %>% kable() %>% kable_styling()
```

RRMSE:
```{r}
tmp = c(rrmse(simdata, m.I), rrmse(simdata, m.V), rrmse(simdata, m.mV), rrmse(simdata, m.true))
barplot(tmp, names.arg = c('I only', 'V only', 'mV', 'true'))
```

## Data with signals

```{r}
set.seed(2018)
B1 = matrix(0, n/2, p)
V.1 = matrix(0,p,p); V.1[1,1] = 1
B2 = rmvnorm(n/2, sigma = V.1)
B = rbind(B1, B2)

V.random = array(0, dim=c(p,p,n))
ind = sample(1:n, n/2)
V.random[,,ind] = V
V.random[,,-ind] = diag(p)

Ehat = matrix(0, n, p)
Ehat[ind,] = rmvnorm(n/2, sigma = V)
Ehat[-ind,] = rmvnorm(n/2, sigma = diag(p))

Bhat = B + Ehat
simdata = list(B = B, Bhat=Bhat, Shat = 1)
```

```{r}
data = mash_set_data(Bhat, Shat=1)
U.c = cov_canonical(data)
m.I = mash(data, U.c, verbose = FALSE)

data.V = mash_update_data(data, V = V)
m.V = mash(data.V, U.c, verbose = FALSE)

V.mV = estimate_null_correlation(data, U.c, details=TRUE)
m.mV = V.mV$mash.model

data.true = mash_update_data(data, V = V.random)
m.true = mash(data.true, U.c, verbose = FALSE, algorithm.version = 'R')
```

The estimated V is
```{r}
V.mV$V
```

```{r}
Ionly = c(get_loglik(m.I), length(get_significant_results(m.I)), sum(get_significant_results(m.I) <= n/2))

Vonly = c(get_loglik(m.V), length(get_significant_results(m.V)), sum(get_significant_results(m.V) <= n/2))

mV = c(get_loglik(m.mV), length(get_significant_results(m.mV)), sum(get_significant_results(m.mV) <= n/2))

true = c(get_loglik(m.true), length(get_significant_results(m.true)), sum(get_significant_results(m.true) <= n/2))

tmp = rbind(Ionly, Vonly, mV, true)
row.names(tmp) = c('I only', 'V only', 'mV', 'true')
colnames(tmp) = c('loglik', '# signif', 'false positive')
tmp %>% kable() %>% kable_styling()
```

RRMSE:
```{r}
tmp = c(rrmse(simdata, m.I), rrmse(simdata, m.V), rrmse(simdata, m.mV), rrmse(simdata, m.true))
barplot(tmp, names.arg = c('I only', 'V only', 'mV', 'true'))
```

ROC:
```{r}
roc.seq = ROC.table(simdata$B, m.true)
plot(roc.seq[,'FPR'], roc.seq[,'TPR'], type='l', xlab = 'FPR', ylab='TPR',
       main=paste0(' True Pos vs False Pos'), cex=1.5, lwd = 1.5, col = 'cyan')
roc.seq = ROC.table(simdata$B, m.mV)
lines(roc.seq[,'FPR'], roc.seq[,'TPR'], col='purple', lwd = 1.5)
roc.seq = ROC.table(simdata$B, m.I)
lines(roc.seq[,'FPR'], roc.seq[,'TPR'], col='red', lwd = 1.5)
roc.seq = ROC.table(simdata$B, m.V)
lines(roc.seq[,'FPR'], roc.seq[,'TPR'], col='darkolivegreen4', lwd = 1.5)
legend('bottomright', c('oracle','mV', 'I only', 'V only'), col=c('cyan','purple','red','darkolivegreen4'),
           lty=c(1,1,1,1), lwd=c(1.5,1.5,1.5,1.5))
```
