---
title: "Estimate cor — MLE EM V"
author: "Yuxin Zou"
date: 2018-8-3
output: 
  workflowr::wflow_html:
    code_folding: hide
---

```{r}
library(mashr)
library(knitr)
library(kableExtra)
source('../code/estimate_cor.R')
source('../code/generateDataV.R')
source('../code/summary.R')
```

## EM

We use EM algorithm to update V.

B is the $n\times p$ true value matrix. $\mathbf{z}$ is a length n vector.

### E step

$$
P(\hat{B},B,\mathbf{z}|\rho, \pi) = \prod_{i=1}^{n} \prod_{k=0}^{K}\left[\pi_{k}N(\hat{b}_{i}; b_{i}, V)N(b_{i}; 0, U_{k})\right]^{\mathbb{I}(z_{i}=k)}
$$

$$
\mathbb{E}_{\mathbf{z},B|\hat{B}} \log P(\hat{B},B,\mathbf{z}|\rho, \pi) = \sum_{i=1}^{n} \sum_{k=0}^{K} P(z_{i}=k|\hat{b}_{i})\left[ \log \pi_{k} + \mathbb{E}_{B|\hat{B}}(\log N(\hat{b}_{i}; b_{i}, V)) + \mathbb{E}_{B|\hat{B}}(\log N(b_{i}; 0, U_{k})) \right]
$$

$$
\begin{align*}
\log N(\hat{b}_{i}; b_{i}, V) + \log N(b_{i}; 0, U_{k}) &= -\frac{p}{2}\log 2\pi -\frac{1}{2}\log |V| - \frac{1}{2}(\hat{b}_{i}-b_{i})^{T}V^{-1}(\hat{b}_{i}-b_{i}) -\frac{p}{2}\log 2\pi -\frac{1}{2}\log |U_{k}| - \frac{1}{2}b_{i}^{T}U_{k}^{-1}b_{i} \\
&= -p\log 2\pi -\frac{1}{2}\log |U_{k}| -\frac{1}{2}\log |V| - \frac{1}{2}\hat{b}_{i}^{T}V^{-1}\hat{b}_{i} + \hat{b}_{i}^{T}V^{-1}b_{i} -\frac{1}{2}b_{i}^{T}V^{-1}b_{i} - \frac{1}{2}b_{i}^{T}U_{k}^{-1}b_{i} \\
\mathbb{E}_{b_{i}|\hat{b}_{i}}\left[ \log N(\hat{b}_{i}; b_{i}, V) + \log N(b_{i}; 0, U_{k}) \right] &= -p\log 2\pi -\frac{1}{2}\log |U_{k}| -\frac{1}{2}\log |V| - \frac{1}{2}\hat{b}_{i}^{T}V^{-1}\hat{b}_{i} + \hat{b}_{i}^{T}V^{-1}\mathbb{E}(b_{i}|\hat{b}_{i}) -\frac{1}{2}tr\left(V^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right) - \frac{1}{2}tr\left(U_{k}^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right)
\end{align*}
$$

$$
\gamma_{z_{i}}(k) = P(z_{i}=k|X_{i}) = \frac{\pi_{k}N(x_{i}; 0, V+U_{k})}{\sum_{k'=0}^{K}\pi_{k'}N(x_{i}; 0, V + U_{k'})}
$$

### M step

$V$:
$$
\begin{align*}
f(V^{-1}) = \sum_{i=1}^{n} \sum_{k=0}^{K} \gamma_{Z_{i}}(k)\left[ -p\log 2\pi -\frac{1}{2}\log |U_{k}| -\frac{1}{2}\log |V| - \frac{1}{2}\hat{b}_{i}^{T}V^{-1}\hat{b}_{i} + \hat{b}_{i}^{T}V^{-1}\mathbb{E}(b_{i}|\hat{b}_{i}) -\frac{1}{2}tr\left(V^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right) - \frac{1}{2}tr\left(U_{k}^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right) \right]
\end{align*}
$$

$$
\begin{align*}
f(V^{-1})' &= \sum_{i=1}^{n} \sum_{k=0}^{K} \gamma_{Z_{i}}(k)\left[ \frac{1}{2}V - \frac{1}{2}\hat{b}_{i}\hat{b}_{i}^{T} + \mathbb{E}(b_{i}|\hat{b}_{i})\hat{b}_{i}^{T} - \frac{1}{2} \mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right] = 0 \\
\frac{1}{2}Vn &= \sum_{i=1}^{n} \sum_{k=0}^{K} \gamma_{Z_{i}}(k)\left[\frac{1}{2}\hat{b}_{i}\hat{b}_{i}^{T} - \mathbb{E}(b_{i}|\hat{b}_{i})\hat{b}_{i}^{T} + \frac{1}{2} \mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right] \\
\hat{V} &= \frac{1}{n} \sum_{i=1}^{n} \left[\hat{b}_{i}\hat{b}_{i}^{T} - 2\mathbb{E}(b_{i}|\hat{b}_{i})\hat{b}_{i}^{T} + \mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right] \\
&= \frac{1}{n} \sum_{i=1}^{n} \mathbb{E}\left[ (\hat{b}_{i} - b_{i})(\hat{b}_{i} - b_{i})^{T} | \hat{b}_{i}\right]
\end{align*}
$$

Algorithm: 
```{r, eval=FALSE, tidy=FALSE, highlight=FALSE}
Input: X, Ulist, init_V
Compute loglikelihood
delta = 1
while delta > tol
  Given rho, Estimate pi using convex method (current mash method)
  M step: update V --> cov2cor(V)
  Compute loglikelihood
  Update delta
```

## Data

$$
\hat{\beta}|\beta \sim N_{2}(\hat{\beta}; \beta, \left(\begin{matrix} 1 & 0.5 \\
                                          0.5 & 1 \end{matrix}\right))
$$

$$
\beta \sim \frac{1}{4}\delta_{0} + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 0 \\
                                          0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 0 & 0 \\
                                          0 & 1 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 1 \\
                                          1 & 1 \end{matrix}\right))
$$

n = 4000

```{r}
set.seed(1)
n = 4000; p = 2
Sigma = matrix(c(1,0.5,0.5,1),p,p)
U0 = matrix(0,2,2)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[2,2] = 1
U3 = matrix(1,2,2)
Utrue = list(U0=U0, U1=U1, U2=U2, U3=U3)
data = generate_data(n, p, Sigma, Utrue)
```

```{r}
m.data = mash_set_data(data$Bhat, data$Shat)
U.c = cov_canonical(m.data)
grid = mashr:::autoselect_grid(m.data, sqrt(2))
Ulist = mashr:::normalize_Ulist(U.c)
xUlist = mashr:::expand_cov(Ulist,grid,usepointmass =  TRUE)
result <- mixture.EMV.times(data$Bhat, xUlist, init_V = list(diag(ncol(data$Bhat)), matrix(c(1,-0.8,-0.8,1),2,2)), grid=1)
plot(result$result$log_liks)
```
The estimated V is 
```{r}
result$result$V
```

```{r}
m.data.em = mash_set_data(data$Bhat, data$Shat, V = result$result$V)
U.c = cov_canonical(m.data.em)
m.em = mash(m.data.em, U.c, verbose= FALSE)
null.ind = which(apply(data$B,1,sum) == 0)
```

The log likelihood is `r round(get_loglik(m.em),2)`. There are `r length(get_significant_results(m.em))` significant samples, `r sum(get_significant_results(m.em) %in% null.ind)` false positives. The RRMSE is `r RRMSE(data$B, data$Bhat, list(m.em))`.

The estimated `pi` is
```{r}
barplot(get_estimated_pi(m.em), las=2, cex.names = 0.7, main='EM V', ylim=c(0,0.8))
```


The ROC curve:
```{r}
m.data.correct = mash_set_data(data$Bhat, data$Shat, V=Sigma)
m.correct = mash(m.data.correct, U.c, verbose = FALSE)
m.correct.seq = ROC.table(data$B, m.correct)
m.em.seq = ROC.table(data$B, m.em)
```

```{r echo=FALSE, fig.align = "center"}
{plot(m.correct.seq[,'FPR'], m.correct.seq[,'TPR'],type='l',xlab = 'FPR', ylab='TPR', main='True Positive vs False Positive', cex=1.5, lwd = 1.5)
lines(m.em.seq[,'FPR'], m.em.seq[,'TPR'], col='red', lwd = 1.5)
legend('bottomright', c('True','EM_V'),col=c('black','red'),lty=c(1,1), lwd=c(1.5,1.5))}
```
