---
title: "Estimate cor—max MASH"
author: "Yuxin Zou"
date: 2018-07-25
output: 
  html_document:
    code_folding: hide
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

```{r}
library(mashr)
source('../code/generateDataV.R')
source('../code/estimate_cor.R')
library(knitr)
library(kableExtra)
```

# One example
$$
\hat{\beta}|\beta \sim N_{3}(\hat{\beta}; \beta, \left(\begin{matrix} 1 & 0.7 & 0.1 \\
                                          0.7 & 1 & 0.5 \\ 
                                          0.1 & 0.5 & 1 \end{matrix}\right))
$$

$$
\beta \sim \frac{1}{4}\delta_{0} + \frac{1}{4}N_{3}(0, \left(\begin{matrix} 1 & 0 &0\\
                                          0 & 0 & 0 \\
                                          0 & 0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{3}(0, \left(\begin{matrix} 1 & 0 & 0 \\
                     0 & 1 & 0 \\
                     0 & 0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{3}(0, \left(\begin{matrix} 1 & 1 & 1 \\
                     1 & 1 & 1 \\
                     1 & 1 & 1 \end{matrix}\right))
$$

```{r}
set.seed(1)
Sigma = cbind(c(1,0.7,0.1), c(0.7,1,0.5), c(0.1,0.5,1))
U0 = matrix(0,3,3)
U1 = matrix(0,3,3); U1[1,1] = 1
U2 = diag(3); U2[3,3] = 0
U3 = matrix(1,3,3)
data = generate_data(n=500, p=3, V=Sigma, Utrue = list(U0=U0, U1=U1,U2=U2,U3=U3))
```

```{r}
m.data = mash_set_data(data$Bhat, data$Shat)
m.1by1 = mash_1by1(m.data)
strong = get_significant_results(m.1by1)

U.pca = cov_pca(m.data, 3, subset = strong)
U.ed = cov_ed(m.data, U.pca, subset = strong)
U.c = cov_canonical(m.data)
```

We run the algorithm in [estimate cor mle](EstimateCorMax.html) with 3 different initial points for $\rho$. The estimated correlation is

```{r, warning=FALSE}
Vhat.max = estimateV(m.data, c(U.c, U.ed), init_rho = c(-0.5,0,0.5), tol=1e-3)
Vhat.max$V
```

The running time for each pairwise correlation is 
```{r}
table = data.frame(t(Vhat.max$ttime))
colnames(table) = c('12','13','23')
table %>% kable() %>% kable_styling()
```
The time is the total running time with different initial point.

Using the original truncated correlation:
```{r}
Vhat.tru = estimate_null_correlation(m.data)
Vhat.tru
```

Checking the estiamted error:
```{r}
maxError = c(max(abs(Vhat.max$V - Sigma)),
max(abs(Vhat.tru - Sigma)))
OpError = c(svd(Vhat.max$V - Sigma)$d[1],
svd(Vhat.tru - Sigma)$d[1])
table = data.frame(maxError = maxError, OpError = OpError, row.names = c('mle','trunc'))
table %>% kable() %>% kable_styling()
```
In mash model, the model with correlation from mle has larger loglikelihood.

```{r}
m.data.mle = mash_set_data(data$Bhat, data$Shat, V=Vhat.max$V)
m.model.mle = mash(m.data.mle, c(U.c,U.ed), verbose = FALSE)
```

```{r}
m.data.trunc = mash_set_data(data$Bhat, data$Shat, V=Vhat.tru)
m.model.trunc = mash(m.data.trunc, c(U.c,U.ed), verbose = FALSE)
```

```{r}
table = data.frame(log_likelihood = c(get_loglik(m.model.mle), get_loglik(m.model.trunc)), row.names = c('mle', 'trunc'))
t(table) %>% kable() %>% kable_styling()
```

# More simulations

I randomly generate 10 positive definite correlation matrices, V. The sample size is 2000.

$$
\hat{z}|z \sim N_{5}(z, V)
$$
$$
z\sim\frac{1}{4}\delta_{0} + \frac{1}{4}N_{5}(0,\left(\begin{matrix} 1 & \mathbf{0}_{1\times 4} \\ \mathbf{0}_{4\times 1} & \mathbf{0}_{4\times 4} \end{matrix}\right)) + \frac{1}{4}N_{5}(0,\left(\begin{matrix} \mathbf{1}_{2\times 2} & \mathbf{0}_{1\times 3} \\ \mathbf{0}_{3\times 1} & \mathbf{0}_{3\times 3} \end{matrix}\right)) + \frac{1}{4}N_{5}(0,\mathbf{1}_{5\times 5})
$$

```{r, warning=FALSE, eval=FALSE}
library(mashr)
n=500; p = 5
U0 = matrix(0,p,p)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[c(1:2), c(1:2)] = 1
U3 = matrix(1, p,p)
Utrue = list(U0 = U0, U1 = U1, U2 = U2, U3 = U3)
result = V.simulation(n,p,10,Utrue)
saveRDS(result, '../output/MASH.10.mle.result.rds')
```
```{r}
result = readRDS('../output/MASH.10.mle.result.rds')
```

The max elementwise norm is
```{r}
temp = rbind(result$mMaxE.max, result$mMaxE.trun)
barplot(temp, axes=T,
        main="max elementwise error",
        names.arg = 1:10,
        legend.text = c("MLE", "Trunc"),
        ylim = c(0, max(colSums(temp))*2))
```

The spectral norm is
```{r}
temp = rbind(result$m2E.max, result$m2E.trun)
barplot(temp, axes=T,
        main="Matrix 2 norm error",
        names.arg = 1:10,
        legend.text = c("MLE", "Trunc"),
        ylim = c(0, max(colSums(temp))*2))
```

The 10 estimated correlation matrices are all positive definite.

The total running time for each matrix is
```{r}
t(result$Time) %>% kable() %>% kable_styling()
```

# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
