---
title: "Estimate cor — EM 1"
author: "Yuxin Zou"
date: 2018-8-2
output: 
  html_document:
    code_folding: hide
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

```{r}
library(mashr)
source('../code/generateDataV.R')
source('../code/summary.R')
```

# EM

## E step

$$
P(X,\mathbf{z}|\rho, \pi) = \prod_{i=1}^{n} \prod_{k=0}^{K}\left[\pi_{k}N(x_{i}; 0, \Sigma_{k})\right]^{\mathbb{I}(z_{i}=k)} \prod_{k=0}^{K}\pi_{k}^{\lambda_{k}-1}
$$

$$
\mathbb{E}_{\mathbf{z}|X} \log P(X,\mathbf{z}|\rho, \pi) = \sum_{i=1}^{n} \sum_{k=0}^{K} P(z_{i}=k|X)\left[ \log \pi_{k} + \log N(x_{i}; 0, \Sigma_{k})\right] + \sum_{k=0}^{K} (\lambda_{k}-1)\log \pi_{k}
$$

$$
\gamma_{z_{i}}(k) = P(z_{i}=k|X_{i}) = \frac{\pi_{k}N(x_{i}; 0, \Sigma_{k})}{\sum_{k'=0}^{K}\pi_{k'}N(x_{i}; 0, \Sigma_{k'})}
$$

## M step

$\pi$: 
$$
\sum_{i=1}^{n} \gamma_{z_{i}}(k) \frac{1}{\pi_{k}} + \frac{\lambda_{k}-1}{\pi_{k}} - \lambda = 0 \quad \rightarrow \pi_{k} = \frac{1}{\lambda} \left(\sum_{i=1}^{n} \gamma_{z_{i}}(k) + \lambda_{k}-1\right) \quad \lambda = n + \sum_{k=1}^{K}\lambda_{k} - K 
$$

$$
\hat{\pi}_{k} = \frac{\sum_{i=1}^{n} \gamma_{z_{i}}(k) + \lambda_{k} - 1 }{n + \sum_{k=1}^{K}\lambda_{k} - K } 
$$

$\rho$:
$$
\begin{align*}
f(\rho) &= \sum_{i=1}^{n} \sum_{k=1}^{K} \gamma_{z_{i}}(k)\left[ -\frac{1}{2}\log (1-\phi_{k}^2)-\frac{1}{2(1-\phi_{k}^2)}\left[ \frac{x_{i}^2}{\sigma_{k11}^2} + \frac{y_{i}^2}{\sigma_{k22}^2} - \frac{2\phi_{k}x_{i}y_{i}}{\sigma_{k11}\sigma_{k22}}\right] \right]\\
f(\rho)' &= \sum_{i=1}^{n} \sum_{k=1}^{K} \gamma_{z_{i}}(k)\left[ \frac{\phi_{k}}{1-\phi_{k}^2}-\frac{\phi_{k}}{(1-\phi_{k}^2)^2}\left[ \frac{x_{i}^2}{\sigma_{k11}^2} + \frac{y_{i}^2}{\sigma_{k22}^2}\right] - \frac{\phi_{k}+1}{(1-\phi_{k}^2)^2}\frac{x_{i}y_{i}}{\sigma_{k11}\sigma_{k22}}\right]\frac{1}{\sigma_{k11}\sigma_{k22}} = 0
\end{align*}
$$
$\phi_k = \frac{\rho + u_{k12}}{\sigma_{k11}\sigma_{k22}}$, $\phi_{k}$ is a function of $\rho$.

Algorithm: 
```{r, eval=FALSE, tindy = FALSE}
Input: X, Ulist, init_rho, init_pi
Compute loglikelihood
delta = 1
while delta > tol
  E step: update z
  M step: update pi, update rho
  Compute loglikelihood
  Update delta
```

```{r}
get_sigma <- function(rho, Ulist){
  V <- matrix(c(1,rho,rho,1), 2,2)
  lapply(Ulist, function(U) U + V)
}

penalty <- function(prior, pi_s){
  subset <- (prior != 1.0)
  sum((prior-1)[subset]*log(pi_s[subset]))
}

compute.log.lik <- function(lL, p, prior){
  p = normalize(pmax(0,p))
  temp = log(exp(lL$loglik_matrix) %*% p)+lL$lfactors
  return(sum(temp) + penalty(prior, p))
  # return(sum(temp))
}

normalize <- function(x){
  x/sum(x)
}
```

```{r}
mixture.EM.times <- function(X, Ulist, init_rho=0, init_pi=NULL, prior = c('nullbiased', 'uniform'), control = list()){
  times = length(init_rho)
  result = list()
  loglik = c()
  rho = c()
  time.t = c()
  converge.status = c()
  for(i in 1:times){
    out.time = system.time(result[[i]] <- mixture.EM(X, Ulist,
                                                     init_pi=init_pi,
                                                     init_rho=init_rho[i],
                                                     prior=prior, 
                                                     control = control))
    time.t = c(time.t, out.time['elapsed'])
    loglik = c(loglik, -result[[i]]$B)
    rho = c(rho, result[[i]]$rhohat)
    converge.status = c(converge.status, result[[i]]$converged)
  }
  if(abs(max(loglik) - min(loglik)) < 1e-4){
    status = 'global'
  }else{
    status = 'local'
  }
  ind = which.max(loglik)
  return(list(result = result[[ind]], status = status, loglik = loglik, rho=rho, time = time.t, converge.status = converge.status))
}

mixture.EM <- function(X, Ulist, init_rho=0, init_pi = NULL, prior = c('nullbiased', 'uniform'), control = list()) {
  prior = match.arg(prior)
  prior <- mashr:::set_prior(length(Ulist), prior)
  k = length(Ulist)
  if (is.null(init_pi)){
    init_pi <- rep(1/k,k)
  }
  control = ashr:::set_control_squarem(control,nrow(X))
  res = SQUAREM::squarem(par=c(init_pi, init_rho),fixptfn=fixpoint, objfn=negpenloglik,X=X, Ulist=Ulist, prior=prior, control=control)
  
  return(list(pihat = normalize(pmax(0,head(res$par, -1))), rho = tail(res$par, 1), loglik=-res$value.objfn, niter = res$iter, converged=res$convergence, control=control))
}

fixpoint = function(para, X, Ulist, prior){
  rho = tail(para,1)
  pi_s = head(para, -1)
  pi_s = normalize(pmax(0,pi_s)) #avoid occasional problems with negative pis due to rounding
  
  # compute L
  Sigma <- get_sigma(rho, Ulist)
  L <- t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U)}))
  
  # E
  m  = t(pi_s * t(L)) # matrix_lik is n by k; so this is also n by k
  m.rowsum = rowSums(m)
  classprob = m/m.rowsum #an n by k matrix
  
  # M
  pinew = normalize(colSums(classprob) + prior - 1)
  
  rhonew = optimize(EMloglikelihood, interval = c(-1,1), maximum = TRUE, X = X, Ulist = Ulist, z = classprob)$maximum
  
  return(c(pinew,rhonew))
}

EMloglikelihood = function(rho, X, Ulist, z){
  Sigma = get_sigma(rho, Ulist)
  L = t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U, log=TRUE)}))
  sum(L * z)
}

negpenloglik = function(para, X, Ulist, prior){
  Sigma <- get_sigma(tail(para,1), Ulist)
  lL <- t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U, log=TRUE)}))
  lfactors <- apply(lL,1,max)
  matrix_llik <- lL - lfactors
  lL = list(loglik_matrix = matrix_llik,
              lfactors   = lfactors)
  ll <- compute.log.lik(lL, head(para, -1), prior)
  
  return(-ll)
}
```

## Data

$$
\hat{\beta}|\beta \sim N_{2}(\hat{\beta}; \beta, \left(\begin{matrix} 1 & 0.5 \\
                                          0.5 & 1 \end{matrix}\right))
$$

$$
\beta \sim \frac{1}{4}\delta_{0} + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 0 \\
                                          0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 0 & 0 \\
                                          0 & 1 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 1 \\
                                          1 & 1 \end{matrix}\right))
$$

n = 4000

```{r}
set.seed(1)
n = 4000; p = 2
Sigma = matrix(c(1,0.5,0.5,1),p,p)
U0 = matrix(0,2,2)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[2,2] = 1
U3 = matrix(1,2,2)
Utrue = list(U0=U0, U1=U1, U2=U2, U3=U3)
data = generate_data(n, p, Sigma, Utrue)
```

```{r}
m.data = mash_set_data(data$Bhat, data$Shat)
U.c = cov_canonical(m.data)
grid = mashr:::autoselect_grid(m.data, sqrt(2))
Ulist = mashr:::normalize_Ulist(U.c)
xUlist = mashr:::expand_cov(Ulist,grid,usepointmass =  TRUE)

result.em <- mixture.EM(m.data$Bhat, xUlist)
```
The estimated $\rho$ is `r result.em$rho`.

```{r}
m.data.em = mash_set_data(data$Bhat, data$Shat, V = matrix(c(1,result.em$rho,result.em$rho,1),2,2))
U.c = cov_canonical(m.data.em)
m.em = mash(m.data.em, U.c, verbose= FALSE)
null.ind = which(apply(data$B,1,sum) == 0)
```

The log likelihood is `r round(get_loglik(m.em),2)`. There are `r length(get_significant_results(m.em))` significant samples, `r sum(get_significant_results(m.em) %in% null.ind)` false positives. The RRMSE is `r RRMSE(data$B, data$Bhat, list(m.em))`.

## Another EM version

Algorithm: 
```{r, eval=FALSE, tindy = FALSE}
Input: X, Ulist, init_rho
Given rho, estimate pi by max loglikelihood (convex problem)
Compute loglikelihood
delta = 1
while delta > tol
  E step: compute z
  M step: update rho
  Given rho, estimate pi by max loglikelihood (convex problem)
  Compute loglikelihood
  Update delta
```

```{r}
mixture.M.times <- function(X, Ulist, init_rho=0, init_pi=NULL, prior = c('nullbiased', 'uniform'), control = list()){
  times = length(init_rho)
  result = list()
  loglik = c()
  rho = c()
  time.t = c()
  converge.status = c()
  for(i in 1:times){
    out.time = system.time(result[[i]] <- mixture.M(X, Ulist,
                                                     init_pi=init_pi,
                                                     init_rho=init_rho[i],
                                                     prior=prior, 
                                                     control = control))
    time.t = c(time.t, out.time['elapsed'])
    loglik = c(loglik, -result[[i]]$B)
    rho = c(rho, result[[i]]$rhohat)
    converge.status = c(converge.status, result[[i]]$converged)
  }
  if(abs(max(loglik) - min(loglik)) < 1e-4){
    status = 'global'
  }else{
    status = 'local'
  }
  ind = which.max(loglik)
  return(list(result = result[[ind]], status = status, loglik = loglik, rho=rho, time = time.t, converge.status = converge.status))
}

mixture.M <- function(X, Ulist, init_rho=0, init_pi = NULL, tol=1e-5, prior = c('nullbiased', 'uniform')) {
  prior <- match.arg(prior)

  m.model = fit_mash(X, Ulist, rho = init_rho, prior=prior)
  pi_s = get_estimated_pi(m.model, dimension = 'all')
  # get complete Ulist
  xUlist = mashr:::expand_cov(Ulist, m.model$fitted_g$grid, m.model$fitted_g$usepointmass)
  prior.v <- mashr:::set_prior(length(pi_s), prior)
  
  # compute loglikelihood
  log_liks <- c()
  log_liks <- c(log_liks, get_loglik(m.model)+penalty(prior.v, pi_s))
  delta.ll <- 1
  niter <- 0
  rho = init_rho
  
  while(delta.ll > tol){
    # compute L
    Sigma <- get_sigma(rho, xUlist)
    L <- t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U)}))
  
    m  = t(pi_s * t(L)) # matrix_lik is n by k; so this is also n by k
    m.rowsum = rowSums(m)
    classprob = m/m.rowsum #an n by k matrix
  
    # max_rho
    rho <- optimize(EMloglikelihood, interval = c(-1,1), maximum = TRUE, X = X, Ulist = xUlist, z = classprob)$maximum
    
    m.model = fit_mash(X, Ulist, rho, prior=prior)
    
    pi_s = get_estimated_pi(m.model, dimension = 'all')
    log_liks <- c(log_liks, get_loglik(m.model)+penalty(prior.v, pi_s))
    # Update delta
    delta.ll <- log_liks[length(log_liks)] - log_liks[length(log_liks)-1]
    niter <- niter + 1
  }

  return(list(pihat = normalize(pi_s), rho = rho, loglik=log_liks))
}

EMloglikelihood = function(rho, X, Ulist, z){
  Sigma = get_sigma(rho, Ulist)
  L = t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U, log=TRUE)}))
  sum(L * z)
}

fit_mash <- function(X, Ulist, rho, prior=c('nullbiased', 'uniform')){
  m.data = mashr::mash_set_data(Bhat=X, Shat=1, V = matrix(c(1, rho, rho, 1), 2, 2))
  m.model = mashr::mash(m.data, Ulist, prior=prior, verbose = FALSE)
  return(m.model)
}

```


```{r}
set.seed(1)
n = 4000; p = 2
Sigma = matrix(c(1,0.5,0.5,1),p,p)
U0 = matrix(0,2,2)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[2,2] = 1
U3 = matrix(1,2,2)
Utrue = list(U0=U0, U1=U1, U2=U2, U3=U3)
data = generate_data(n, p, Sigma, Utrue)
```

```{r}
m.data = mash_set_data(data$Bhat, data$Shat)
U.c = cov_canonical(m.data)
# grid = mashr:::autoselect_grid(m.data, sqrt(2))
# Ulist = mashr:::normalize_Ulist(U.c)
# xUlist = mashr:::expand_cov(Ulist,grid,usepointmass=TRUE)

result.m <- mixture.M(m.data$Bhat, Ulist)
```

The estimated $\rho$ is `r result.m$rho`.

```{r}
m.data.m = mash_set_data(data$Bhat, data$Shat, V = matrix(c(1,result.m$rho,result.m$rho,1),2,2))
U.c.m = cov_canonical(m.data.m)
m.m = mash(m.data.m, U.c, verbose= FALSE)
null.ind = which(apply(data$B,1,sum) == 0)
```

The log likelihood is `r round(get_loglik(m.m),2)`. There are `r length(get_significant_results(m.m))` significant samples, `r sum(get_significant_results(m.m) %in% null.ind)` false positives. The RRMSE is `r RRMSE(data$B, data$Bhat, list(m.m))`.

# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```

