---
title: "Estimate cor—max MASH"
author: "Yuxin Zou"
date: 2018-07-25
output: 
  html_document:
    code_folding: hide
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

```{r}
library(mashr)
source('../code/generateDataV.R')
source('../code/estimate_cor.R')
source('../code/summary.R')
library(knitr)
library(kableExtra)
```

Apply the max methods for correlation matrix on mash data. 

The estimated V from MLE(`optim` function) and EM perform better than the truncated correlation (error, mash log likelihood, ROC).

Comparing the estimated V from MLE and EM, EM algorithm tends to compute faster, and the estimated correlation is slightly better than the one from MLE in terms of estimation error, mash log likelihood, ROC.

# One example
$$
\hat{\beta}|\beta \sim N_{3}(\hat{\beta}; \beta, \left(\begin{matrix} 1 & 0.7 & 0.2 \\
                                          0.7 & 1 & 0.4 \\ 
                                          0.2 & 0.4 & 1 \end{matrix}\right))
$$

$$
\beta \sim \frac{1}{4}\delta_{0} + \frac{1}{4}N_{3}(0, \left(\begin{matrix} 1 & 0 &0\\
                                          0 & 0 & 0 \\
                                          0 & 0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{3}(0, \left(\begin{matrix} 1 & 0 & 0 \\
                     0 & 1 & 0 \\
                     0 & 0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{3}(0, \left(\begin{matrix} 1 & 1 & 1 \\
                     1 & 1 & 1 \\
                     1 & 1 & 1 \end{matrix}\right))
$$

```{r}
set.seed(1)
Sigma = cbind(c(1,0.7,0.2), c(0.7,1,0.4), c(0.2,0.4,1))
U0 = matrix(0,3,3)
U1 = matrix(0,3,3); U1[1,1] = 1
U2 = diag(3); U2[3,3] = 0
U3 = matrix(1,3,3)
data = generate_data(n=4000, p=3, V=Sigma, Utrue = list(U0=U0, U1=U1,U2=U2,U3=U3))
```

We find the estimate of V with canonical covariances and the PCA covariances.
```{r}
m.data = mash_set_data(data$Bhat, data$Shat)
m.1by1 = mash_1by1(m.data)
strong = get_significant_results(m.1by1)

U.pca = cov_pca(m.data, 3, subset = strong)
U.ed = cov_ed(m.data, U.pca, subset = strong)
U.c = cov_canonical(m.data)
```

The PCA correlation matrices are:
```{r echo=FALSE, fig.width=8, fig.height=4,fig.align='center'}
plotCor = function(U, title, condition_names = 1:nrow(U)){
  x           <- cov2cor(U)
  x[x > 1]    <- 1
  x[x < -1]   <- -1
  colnames(x) <- condition_names
  rownames(x) <- condition_names
  corrplot::corrplot.mixed(x,upper='color',cl.lim=c(-1,1), upper.col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF",
                               "#E0F3F8","#91BFDB","#4575B4")))(40),
                          title=title,mar=c(0,0,1.5,0))
}
par(mfrow=c(2,2))
plotCor(U.ed$ED_PCA_1, 'PCA1')
plotCor(U.ed$ED_PCA_2, 'PCA2')
plotCor(U.ed$ED_PCA_3, 'PCA3')
plotCor(U.ed$ED_tPCA, 'tPCA')
```

1. We run the algorithm in [estimate cor mle](EstimateCorMax.html) with 3 different initial points for $\rho$ (-0.5,0,0.5). The $\rho$ in each iteration is estimated using `optim` function. The estimated correlation is

```{r, warning=FALSE}
Vhat.mle = estimateV(m.data, c(U.c, U.ed), init_rho = c(-0.5,0,0.5), tol=1e-4, optmethod = 'mle')
Vhat.mle$V
```

2. The result uses algorithm in [estimate cor em](EstimateCorMaxEM2.html). $\rho$ in each iteration is the root of a third degree polynomial.

```{r, warning=FALSE}
Vhat.em = estimateV(m.data, c(U.c, U.ed), init_rho = c(-0.5,0,0.5), tol = 1e-4, optmethod = 'em2')
Vhat.em$V
```

The running time (in sec.) for each pairwise correlation is 
```{r}
table = data.frame(rbind(Vhat.mle$ttime, Vhat.em$ttime), row.names = c('mle', 'em'))
colnames(table) = c('12','13','23')
table %>% kable() %>% kable_styling()
```
The time is the total running time with different initial point.

3. The result uses algorithm in [estimate cor em V](EstimateCorMaxEMV.html).

```{r, warning=FALSE}
Vhat.emV = estimateV(m.data, c(U.c, U.ed), init_V = list(diag(ncol(m.data$Bhat)), clusterGeneration::rcorrmatrix(3), clusterGeneration::rcorrmatrix(3)), tol = 1e-4, optmethod = 'emV')
Vhat.emV$V
```

4. Using the original truncated correlation:
```{r}
Vhat.tru = estimate_null_correlation(m.data)
Vhat.tru
```

The truncated correlation underestimates the correlations.

5. mash 1by1

```{r}
V.mash = cor((data$Bhat/data$Shat)[-strong,])
V.mash
```

## Error

Check the estimation error:
```{r}
FError = c(norm(Vhat.mle$V - Sigma, 'F'),
           norm(Vhat.em$V - Sigma, 'F'),
           norm(Vhat.emV$V - Sigma, 'F'),
           norm(Vhat.tru - Sigma, 'F'),
           norm(V.mash - Sigma, 'F'))
OpError = c(norm(Vhat.mle$V - Sigma, '2'),
           norm(Vhat.em$V - Sigma, '2'),
           norm(Vhat.emV$V - Sigma, '2'),
           norm(Vhat.tru - Sigma, '2'),
           norm(V.mash - Sigma, '2'))
table = data.frame(FrobeniusError = FError, SpectralError = OpError, row.names = c('mle','em','emV','trunc','m.1by1'))
table %>% kable() %>% kable_styling()
```

## mash log likelihood
In mash model, the model with correlation from mle has larger loglikelihood.

```{r}
m.data.mle = mash_set_data(data$Bhat, data$Shat, V=Vhat.mle$V)
m.model.mle = mash(m.data.mle, c(U.c,U.ed), verbose = FALSE)
```

```{r}
m.data.em = mash_set_data(data$Bhat, data$Shat, V=Vhat.em$V)
m.model.em = mash(m.data.em, c(U.c,U.ed), verbose = FALSE)
```

```{r}
m.data.emV = mash_set_data(data$Bhat, data$Shat, V=Vhat.emV$V)
m.model.emV = mash(m.data.emV, c(U.c,U.ed), verbose = FALSE)
```

```{r}
m.data.trunc = mash_set_data(data$Bhat, data$Shat, V=Vhat.tru)
m.model.trunc = mash(m.data.trunc, c(U.c,U.ed), verbose = FALSE)
```

```{r}
m.data.1by1 = mash_set_data(data$Bhat, data$Shat, V=V.mash)
m.model.1by1 = mash(m.data.1by1, c(U.c,U.ed), verbose = FALSE)
```

```{r}
m.data.correct = mash_set_data(data$Bhat, data$Shat, V=Sigma)
m.model.correct = mash(m.data.correct, c(U.c,U.ed), verbose = FALSE)
```

The results are summarized in table:
```{r}
null.ind = which(apply(data$B,1,sum) == 0)
V.trun = c(get_loglik(m.model.trunc), length(get_significant_results(m.model.trunc)), sum(get_significant_results(m.model.trunc) %in% null.ind))
V.mle = c(get_loglik(m.model.mle), length(get_significant_results(m.model.mle)), sum(get_significant_results(m.model.mle) %in% null.ind))
V.em = c(get_loglik(m.model.em), length(get_significant_results(m.model.em)), sum(get_significant_results(m.model.em) %in% null.ind))
V.emV = c(get_loglik(m.model.emV), length(get_significant_results(m.model.emV)), sum(get_significant_results(m.model.emV) %in% null.ind))
V.1by1 = c(get_loglik(m.model.1by1), length(get_significant_results(m.model.1by1)), sum(get_significant_results(m.model.1by1) %in% null.ind))
V.correct = c(get_loglik(m.model.correct), length(get_significant_results(m.model.correct)), sum(get_significant_results(m.model.correct) %in% null.ind))
temp = cbind(V.mle, V.em, V.emV, V.trun, V.1by1, V.correct)
colnames(temp) = c('MLE','EM','EMV', 'Truncate', 'm.1by1', 'True')
row.names(temp) = c('log likelihood', '# significance', '# False positive')
temp %>% kable() %>% kable_styling()
```

The estimated `pi` is
```{r fig.align = "center"}
par(mfrow=c(2,3))
barplot(get_estimated_pi(m.model.mle), las=2, cex.names = 0.7, main='MLE', ylim=c(0,0.8))
barplot(get_estimated_pi(m.model.em), las=2, cex.names = 0.7, main='EM', ylim=c(0,0.8))
barplot(get_estimated_pi(m.model.emV), las=2, cex.names = 0.7, main='EMV', ylim=c(0,0.8))
barplot(get_estimated_pi(m.model.trunc), las=2, cex.names = 0.7, main='Truncate', ylim=c(0,0.8))
barplot(get_estimated_pi(m.model.1by1), las=2, cex.names = 0.7, main='m.1by1', ylim=c(0,0.8))
barplot(get_estimated_pi(m.model.correct), las=2, cex.names = 0.7, main='True', ylim=c(0,0.8))
```

## ROC

```{r}
m.mle.seq = ROC.table(data$B, m.model.mle)
m.em.seq = ROC.table(data$B, m.model.em)
m.emV.seq = ROC.table(data$B, m.model.emV)
m.trun.seq = ROC.table(data$B, m.model.trunc)
m.1by1.seq = ROC.table(data$B, m.model.1by1)
m.correct.seq = ROC.table(data$B, m.model.correct)
```

```{r echo=FALSE, fig.align = "center"}
{plot(m.correct.seq[,'FPR'], m.correct.seq[,'TPR'],type='l',xlab = 'FPR', ylab='TPR', main='True Positive vs False Positive', cex=1.5, lwd = 1.5)
lines(m.mle.seq[,'FPR'], m.mle.seq[,'TPR'], col='red', lwd = 1.5)
lines(m.em.seq[,'FPR'], m.em.seq[,'TPR'], col='green', lwd = 1.5)
lines(m.emV.seq[,'FPR'], m.emV.seq[,'TPR'], col='purple', lwd = 1.5)
lines(m.trun.seq[,'FPR'], m.trun.seq[,'TPR'], col='blue', lwd = 1.5)
lines(m.1by1.seq[,'FPR'], m.1by1.seq[,'TPR'], col='cyan', lwd = 1.5)
legend('bottomright', c('True','MLE','EM','EMV','Trunc', 'm.1by1'),col=c('black','red','green','purple', 'blue', 'cyan'),lty=c(1,1,1,1,1,1), lwd=c(1.5,1.5,1.5,1.5,1.5, 1.5))}
```

## RRMSE

```{r}
rrmse = rbind(RRMSE(data$B, data$Bhat, list(m.model.mle, m.model.em, m.model.emV, m.model.trunc, m.model.1by1, m.model.correct)))
colnames(rrmse) = c('MLE','EM','EMV', 'Truncate','m.1by1','True')
row.names(rrmse) = 'RRMSE'
rrmse %>% kable() %>% kable_styling()
```

```{r}
barplot(rrmse, ylim=c(0,(1+max(rrmse))/2), names.arg = c('MLE','EM', 'EMV','Truncate','m.1by1','True'), las=2, cex.names = 0.7, main='RRMSE')
```

# More simulations

I randomly generate 10 positive definite correlation matrices, V. The sample size is 4000.

$$
\hat{z}|z \sim N_{5}(z, V)
$$
$$
z\sim\frac{1}{4}\delta_{0} + \frac{1}{4}N_{5}(0,\left(\begin{matrix} 1 & \mathbf{0}_{1\times 4} \\ \mathbf{0}_{4\times 1} & \mathbf{0}_{4\times 4} \end{matrix}\right)) + \frac{1}{4}N_{5}(0,\left(\begin{matrix} \mathbf{1}_{2\times 2} & \mathbf{0}_{1\times 3} \\ \mathbf{0}_{3\times 1} & \mathbf{0}_{3\times 3} \end{matrix}\right)) + \frac{1}{4}N_{5}(0,\mathbf{1}_{5\times 5})
$$

```{r, warning=FALSE, eval=FALSE}
set.seed(100)
n=4000; p = 5
U0 = matrix(0,p,p)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[c(1:2), c(1:2)] = 1
U3 = matrix(1, p,p)
Utrue = list(U0 = U0, U1 = U1, U2 = U2, U3 = U3)
for(t in 1:10){
  Vtrue = clusterGeneration::rcorrmatrix(p)
  data = generate_data(n, p, Vtrue, Utrue)
  # mash cov
  m.data = mash_set_data(Bhat = data$Bhat, Shat = data$Shat)
  m.1by1 = mash_1by1(m.data)
  strong = get_significant_results(m.1by1)

  U.pca = cov_pca(m.data, 3, subset = strong)
  U.ed = cov_ed(m.data, U.pca, subset = strong)
  U.c = cov_canonical(m.data)
  Vhat.mle <- estimateV(m.data, c(U.c, U.ed), init_rho = c(-0.5,0,0.5), tol=1e-4, optmethod = 'mle')
  Vhat.em <- estimateV(m.data, c(U.c, U.ed), init_rho = c(-0.5,0,0.5), tol=1e-4, optmethod = 'em2')
  Vhat.emV <- estimateV(m.data, c(U.c, U.ed), init_V = list(diag(ncol(m.data$Bhat)), clusterGeneration::rcorrmatrix(p), clusterGeneration::rcorrmatrix(p)),tol=1e-4, optmethod = 'emV')
  saveRDS(list(V.true = Vtrue, V.mle = Vhat.mle, V.em = Vhat.em, V.emV = Vhat.emV, data = data, strong=strong),
          paste0('../output/MASH.result.',t,'.rds'))
}

```

```{r}
files = dir("../output/AddEMV/"); files = files[grep("MASH.result",files)]
times = length(files)
result = vector(mode="list",length = times)
for(i in 1:times) {
  result[[i]] = readRDS(paste("../output/AddEMV/", files[[i]], sep=""))
}
```

```{r warning=FALSE}
mle.pd = numeric(times)
em.pd = numeric(times)
for(i in 1:times){
  m.data = mash_set_data(result[[i]]$data$Bhat, result[[i]]$data$Shat)
  
  result[[i]]$V.trun = estimate_null_correlation(m.data, apply_lower_bound = FALSE)
  m.1by1 = mash_1by1(m.data)
  strong = get_significant_results(m.1by1)
  result[[i]]$V.1by1 = cor(m.data$Bhat[-strong,])
  U.c = cov_canonical(m.data)
  U.pca = cov_pca(m.data, 3, subset = strong)
  U.ed = cov_ed(m.data, U.pca, subset = strong)

  m.data.true = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = result[[i]]$V.true)
  m.model.true = mash(m.data.true, c(U.c,U.ed), verbose = FALSE)
  
  m.data.trunc = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = result[[i]]$V.trun)
  m.model.trunc = mash(m.data.trunc, c(U.c,U.ed), verbose = FALSE)
  
  m.data.1by1 = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = result[[i]]$V.1by1)
  m.model.1by1 = mash(m.data.1by1, c(U.c,U.ed), verbose = FALSE)
  
  m.data.emV = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = result[[i]]$V.emV$V)
  m.model.emV = mash(m.data.emV, c(U.c,U.ed), verbose = FALSE)
  
  # MLE
  
  m.model.mle = m.model.mle.F = m.model.mle.2 = list()
  R <- tryCatch(chol(result[[i]]$V.mle$V),error = function (e) FALSE)
  if(is.matrix(R)){
    mle.pd[i] = 1
    m.data.mle = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = result[[i]]$V.mle$V)
    m.model.mle = mash(m.data.mle, c(U.c,U.ed), verbose = FALSE)
  }else{
    V.mle.near.F = as.matrix(Matrix::nearPD(result[[i]]$V.mle$V, conv.norm.type = 'F', keepDiag = TRUE)$mat)
    V.mle.near.2 = as.matrix(Matrix::nearPD(result[[i]]$V.mle$V, conv.norm.type = '2', keepDiag = TRUE)$mat)
    
    result[[i]]$V.mle.F = V.mle.near.F
    result[[i]]$V.mle.2 = V.mle.near.2
    
    # mashmodel
    m.data.mle.F = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = V.mle.near.F)
    m.model.mle.F = mash(m.data.mle.F, c(U.c,U.ed), verbose = FALSE)
    
    m.data.mle.2 = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = V.mle.near.2)
    m.model.mle.2 = mash(m.data.mle.2, c(U.c,U.ed), verbose = FALSE)
  }
  
  # EM
  m.model.em = m.model.em.F = m.model.em.2 = list()
  R <- tryCatch(chol(result[[i]]$V.em$V),error = function (e) FALSE)
  if(is.matrix(R)){
    em.pd[i] = 1
    m.data.em = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = result[[i]]$V.em$V)
    m.model.em = mash(m.data.em, c(U.c,U.ed), verbose = FALSE)
  }else{
    V.em.near.F = as.matrix(Matrix::nearPD(result[[i]]$V.em$V, conv.norm.type = 'F', keepDiag = TRUE)$mat)
    V.em.near.2 = as.matrix(Matrix::nearPD(result[[i]]$V.em$V, conv.norm.type = '2', keepDiag = TRUE)$mat)
    
    result[[i]]$V.em.F = V.em.near.F
    result[[i]]$V.em.2 = V.em.near.2
    
    # mashmodel
    m.data.em.F = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = V.em.near.F)
    m.model.em.F = mash(m.data.em.F, c(U.c,U.ed), verbose = FALSE)
      
    m.data.em.2 = mash_set_data(Bhat = m.data$Bhat, Shat = m.data$Shat, V = V.em.near.2)
    m.model.em.2 = mash(m.data.em.2, c(U.c,U.ed), verbose = FALSE)
  }
  
  result[[i]]$m.model = list(m.model.true = m.model.true, m.model.trunc = m.model.trunc, 
                             m.model.1by1 = m.model.1by1, m.model.emV = m.model.emV,
                             m.model.mle = m.model.mle,
                             m.model.mle.F = m.model.mle.F, m.model.mle.2 = m.model.mle.2, 
                             m.model.em = m.model.em,
                             m.model.em.F = m.model.em.F, m.model.em.2 = m.model.em.2)
}
```

## Error

Some estimated correlation matrices are not positive definite. So I estimate the nearest PD cor matrix with `nearPD` function.

The column with `.F`, `.2` are from the nearest positive definite matrix with respect to Frobenius norm and spectral norm.

The Frobenius norm is
```{r}
temp = matrix(0,nrow = times, ncol = 7)
for(i in 1:times){
  temp[i, ] = error.cor(result[[i]], norm.type='F', mle.pd = mle.pd[i], em.pd = em.pd[i])
}
colnames(temp) = c('Trunc','m.1by1', 'MLE','MLE.F', 'EM', 'EM.F', 'EMV')
temp[temp==0] = NA
temp %>% kable() %>% kable_styling()
```

The spectral norm is
```{r}
temp = matrix(0,nrow = times, ncol = 7)
for(i in 1:times){
  temp[i, ] = error.cor(result[[i]], norm.type='2', mle.pd = mle.pd[i], em.pd = em.pd[i])
}
colnames(temp) = c('Trunc','m.1by1', 'MLE','MLE.2', 'EM', 'EM.2', 'EMV')
temp[temp==0] = NA
temp %>% kable() %>% kable_styling()
```

## Time

The total running time for each matrix is
```{r}
mle.time = em.time = numeric(times)
for(i in 1:times){
  mle.time[i] = sum(result[[i]]$V.mle$ttime)
  em.time[i] = sum(result[[i]]$V.em$ttime)
}
temp = cbind(mle.time, em.time)
colnames(temp) = c('MLE', 'EM')
row.names(temp) = 1:10
temp %>% kable() %>% kable_styling()
```

## mash log likelihood

The `NA` means the estimated correlation matrix is not positive definite.

```{r}
temp = matrix(0,nrow = times, ncol = 10)
for(i in 1:times){
  temp[i, ] = loglik.cor(result[[i]]$m.model, mle.pd = mle.pd[i], em.pd = em.pd[i])
}
colnames(temp) = c('True', 'Trunc','m.1by1', 'MLE','MLE.F', 'MLE.2', 'EM', 'EM.F', 'EM.2','EMV')
temp[temp == 0] = NA
temp[,-c(6,9)] %>% kable() %>% kable_styling()
```

## ROC

```{r}
par(mfrow=c(1,2))
for(i in 1:times){
  plotROC(result[[i]]$data$B, result[[i]]$m.model, mle.pd = mle.pd[i], em.pd = em.pd[i], title=paste0('Data', i, ' '))
}
```

## RRMSE

```{r}
par(mfrow=c(1,2))
for(i in 1:times){
  rrmse = rbind(RRMSE(result[[i]]$data$B, result[[i]]$data$Bhat, result[[i]]$m.model))
  barplot(rrmse, ylim=c(0,(1+max(rrmse))/2), las=2, cex.names = 0.7, main='RRMSE')
}
```


# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
