---
title: "Estimate cor — EM 1"
author: "Yuxin Zou"
date: 2018-8-2
output: 
  html_document:
    code_folding: hide
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

```{r}
library(mashr)
source('../code/generateDataV.R')
source('../code/summary.R')
```

# EM

## E step

$$
P(X,\mathbf{z}|\rho, \pi) = \prod_{i=1}^{n} \prod_{p=0}^{P}\left[\pi_{p}N(x_{i}; 0, \Omega_{p})\right]^{\mathbb{I}(z_{i}=p)} \prod_{p=0}^{P}\pi_{p}^{\lambda_{p}-1}
$$

$$
\mathbb{E}_{\mathbf{z}|X} \log P(X,\mathbf{z}|\rho, \pi) = \sum_{i=1}^{n} \sum_{p=0}^{P} P(z_{i}=p|X)\left[ \log \pi_{p} + \log N(x_{i}; 0, \Omega_{p})\right] + \sum_{p=0}^{P} (\lambda_{p}-1)\log \pi_{p}
$$

$$
\gamma_{z_{i}}(p) = P(z_{i}=p|X_{i}) = \frac{\pi_{p}N(x_{i}; 0, \Omega_{p})}{\sum_{p'=0}^{P}\pi_{p'}N(x_{i}; 0, \Omega_{p'})}
$$

## M step

$\pi$: 
$$
\sum_{i=1}^{n} \gamma_{z_{i}}(p) \frac{1}{\pi_{p}} + \frac{\lambda_{p}-1}{\pi_{p}} - \lambda = 0 \quad \rightarrow \pi_{p} = \frac{1}{\lambda} \left(\sum_{i=1}^{n} \gamma_{z_{i}}(p) + \lambda_{p}-1\right) \quad \lambda = n + \sum_{p=1}^{P}\lambda_{p} - P 
$$

$$
\hat{\pi}_{p} = \frac{\sum_{i=1}^{n} \gamma_{z_{i}}(p) + \lambda_{p} - 1 }{n + \sum_{p=1}^{P}\lambda_{p} - P } 
$$

$\rho$:
$$
\begin{align*}
f(\rho) &= \sum_{i=1}^{n} \sum_{p=1}^{P} \gamma_{z_{i}}(p)\left[ -\frac{1}{2}\log (1-\phi_{p}^2)-\frac{1}{2(1-\phi_{p}^2)}\left[ \frac{x_{i}^2}{\omega_{p11}^2} + \frac{y_{i}^2}{\omega_{p22}^2} - \frac{2\phi_{p}x_{i}y_{i}}{\omega_{p11}\omega_{p22}}\right] \right]\\
f(\rho)' &= \sum_{i=1}^{n} \sum_{p=1}^{P} \gamma_{z_{i}}(p)\left[ \frac{\phi_{p}}{1-\phi_{p}^2}-\frac{\phi_{p}}{(1-\phi_{p}^2)^2}\left[ \frac{x_{i}^2}{\omega_{p11}^2} + \frac{y_{i}^2}{\omega_{p22}^2}\right] - \frac{\phi_{p}+1}{(1-\phi_{p}^2)^2}\frac{x_{i}y_{i}}{\omega_{p11}\omega_{p22}}\right]\frac{1}{\omega_{p11}\omega_{p22}} = 0
\end{align*}
$$
$\phi_p = \frac{\rho + \sigma_{p12}}{\omega_{p11}\omega_{p22}}$, $\phi_{p}$ is a function of $\rho$.

Algorithm: 
```{r, eval=FALSE, tindy = FALSE}
Input: X, Ulist, init_rho, init_pi
Compute loglikelihood
delta = 1
while delta > tol
  E step: update z
  M step: update pi, update rho
  Compute loglikelihood
  Update delta
```

```{r}
#' @param rho the off diagonal element of V, 2 by 2 correlation matrix
#' @param Ulist a list of covariance matrices, U_{k}
get_sigma <- function(rho, Ulist){
  V <- matrix(c(1,rho,rho,1), 2,2)
  lapply(Ulist, function(U) U + V)
}

penalty <- function(prior, pi_s){
  subset <- (prior != 1.0)
  sum((prior-1)[subset]*log(pi_s[subset]))
}

#' @title compute log likelihood
#' @param L log likelihoods,
#' where the (i,k)th entry is the log probability of observation i
#' given it came from component k of g
#' @param p the vector of mixture proportions
#' @param prior the weight for the penalty
compute.log.lik <- function(lL, p, prior){
  p = normalize(pmax(0,p))
  temp = log(exp(lL$loglik_matrix) %*% p)+lL$lfactors
  return(sum(temp) + penalty(prior, p))
  # return(sum(temp))
}

normalize <- function(x){
  x/sum(x)
}
```

```{r}
mixture.EM.times <- function(X, Ulist, init_rho=0, init_pi=NULL, prior = c('nullbiased', 'uniform'), control = list()){
  times = length(init_rho)
  result = list()
  loglik = c()
  rho = c()
  time.t = c()
  converge.status = c()
  for(i in 1:times){
    out.time = system.time(result[[i]] <- mixture.EM(X, Ulist,
                                                     init_pi=init_pi,
                                                     init_rho=init_rho[i],
                                                     prior=prior,
                                                     control = control))
    time.t = c(time.t, out.time['elapsed'])
    loglik = c(loglik, result[[i]]$loglik)
    rho = c(rho, result[[i]]$rhohat)
    converge.status = c(converge.status, result[[i]]$converged)
  }
  if(abs(max(loglik) - min(loglik)) < 1e-4){
    status = 'global'
  }else{
    status = 'local'
  }
  ind = which.max(loglik)
  return(list(result = result[[ind]], status = status, loglik = loglik, rho=rho, time = time.t, converge.status = converge.status))
}

mixture.EM <- function(X, Ulist, init_rho=0, init_pi = NULL, prior = c('nullbiased', 'uniform'), control = list()) {
  prior = match.arg(prior)
  prior <- mashr:::set_prior(length(Ulist), prior)
  k = length(Ulist)
  if (is.null(init_pi)){
    init_pi <- rep(1/k,k)
  }
  control = ashr:::set_control_squarem(control,nrow(X))
  res = SQUAREM::squarem(par=c(init_pi, init_rho),fixptfn=fixpoint_EM, objfn=negpenloglik,X=X, Ulist=Ulist, prior=prior, control=control)

  return(list(pihat = normalize(pmax(0,head(res$par, -1))), rhohat = tail(res$par, 1), loglik=-res$value.objfn, niter = res$iter, converged=res$convergence, control=control))
}

fixpoint_EM = function(par, X, Ulist, prior){
  rho = tail(par,1)
  pi_s = head(par, -1)
  pi_s = normalize(pmax(0,pi_s)) #avoid occasional problems with negative pis due to rounding

  # compute L
  Sigma <- get_sigma(rho, Ulist)
  L <- t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U)}))

  # E
  m  = t(pi_s * t(L)) # matrix_lik is n by k; so this is also n by k
  m.rowsum = rowSums(m)
  classprob = m/m.rowsum #an n by k matrix

  # M
  pinew = normalize(colSums(classprob) + prior - 1)

  rhonew = optimize(EMloglikelihood, interval = c(-1,1), maximum = TRUE, X = X, Ulist = Ulist, z = classprob)$maximum

  return(c(pinew,rhonew))
}

EMloglikelihood = function(rho, X, Ulist, z){
  Sigma = get_sigma(rho, Ulist)
  L = t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U, log=TRUE)}))
  sum(L * z)
}

negpenloglik = function(par, X, Ulist, prior){
  Sigma <- get_sigma(tail(par,1), Ulist)
  lL <- t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U, log=TRUE)}))
  lfactors <- apply(lL,1,max)
  matrix_llik <- lL - lfactors
  lL = list(loglik_matrix = matrix_llik,
            lfactors   = lfactors)
  ll <- compute.log.lik(lL, head(par, -1), prior)
  return(-ll)
}
```

# Data

$$
\hat{\beta}|\beta \sim N_{2}(\hat{\beta}; \beta, \left(\begin{matrix} 1 & 0.5 \\
                                          0.5 & 1 \end{matrix}\right))
$$

$$
\beta \sim \frac{1}{4}\delta_{0} + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 0 \\
                                          0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 0 & 0 \\
                                          0 & 1 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 1 \\
                                          1 & 1 \end{matrix}\right))
$$

n = 4000

```{r}
set.seed(1)
n = 4000; p = 2
Sigma = matrix(c(1,0.5,0.5,1),p,p)
U0 = matrix(0,2,2)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[2,2] = 1
U3 = matrix(1,2,2)
Utrue = list(U0=U0, U1=U1, U2=U2, U3=U3)
data = generate_data(n, p, Sigma, Utrue)
```

```{r eval=FALSE}
m.data = mash_set_data(data$Bhat, data$Shat)
U.c = cov_canonical(m.data)
grid = mashr:::autoselect_grid(m.data, sqrt(2))
Ulist = mashr:::normalize_Ulist(U.c)
xUlist = mashr:::expand_cov(Ulist,grid,usepointmass =  TRUE)

result.em <- mixture.EM.times(m.data$Bhat, xUlist)
saveRDS(result.em, '../output/result.em.rds')
```

```{r}
result.em = readRDS('../output/result.em.rds')
```
The estimated $\rho$ is `r result.em$rho`. The running time is `result.em$time` seconds.

```{r}
m.data.em = mash_set_data(data$Bhat, data$Shat, V = matrix(c(1,result.em$rho,result.em$rho,1),2,2))
U.c = cov_canonical(m.data.em)
m.em = mash(m.data.em, U.c, verbose= FALSE)
null.ind = which(apply(data$B,1,sum) == 0)
```

The log likelihood is `r formatC(get_loglik(m.em), digits=7)`. There are `r length(get_significant_results(m.em))` significant samples, `r sum(get_significant_results(m.em) %in% null.ind)` false positives. The RRMSE is `r RRMSE(data$B, data$Bhat, list(m.em))`.

# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```

