---
title: "Estimate cor — MLE EM rho"
author: "Yuxin Zou"
date: 2018-8-3
output: 
  workflowr::wflow_html:
    code_folding: hide
---

```{r}
library(mashr)
library(knitr)
library(kableExtra)
source('../code/estimate_cor.R')
source('../code/generateDataV.R')
source('../code/summary.R')
```

## EM

We use EM algorithm to update $\rho$.

B is the $n\times p$ true value matrix. $\mathbf{z}$ is a length n vector.

### E step

$$
P(\hat{B},B,\mathbf{z}|\rho, \pi) = \prod_{i=1}^{n} \prod_{k=0}^{K}\left[\pi_{k}N(\hat{b}_{i}; b_{i}, V)N(b_{i}; 0, U_{k})\right]^{\mathbb{I}(z_{i}=k)}
$$

$$
\mathbb{E}_{\mathbf{z},B|\hat{B}} \log P(\hat{B},B,\mathbf{z}|\rho, \pi) = \sum_{i=1}^{n} \sum_{k=0}^{K} P(z_{i}=k|\hat{b}_{i})\left[ \log \pi_{k} + \mathbb{E}_{B|\hat{B}}(\log N(\hat{b}_{i}; b_{i}, V)) + \mathbb{E}_{B|\hat{B}}(\log N(b_{i}; 0, U_{k})) \right]
$$

$$
\begin{align*}
\log N(\hat{b}_{i}; b_{i}, V) + \log N(b_{i}; 0, U_{k}) &= -\frac{p}{2}\log 2\pi -\frac{1}{2}\log |V| - \frac{1}{2}(\hat{b}_{i}-b_{i})^{T}V^{-1}(\hat{b}_{i}-b_{i}) -\frac{p}{2}\log 2\pi -\frac{1}{2}\log |U_{k}| - \frac{1}{2}b_{i}^{T}U_{k}^{-1}b_{i} \\
&= -p\log 2\pi -\frac{1}{2}\log |U_{k}| -\frac{1}{2}\log |V| - \frac{1}{2}\hat{b}_{i}^{T}V^{-1}\hat{b}_{i} + \hat{b}_{i}^{T}V^{-1}b_{i} -\frac{1}{2}b_{i}^{T}V^{-1}b_{i} - \frac{1}{2}b_{i}^{T}U_{k}^{-1}b_{i} \\
\mathbb{E}_{b_{i}|\hat{b}_{i}}\left[ \log N(\hat{b}_{i}; b_{i}, V) + \log N(b_{i}; 0, U_{k}) \right] &= -p\log 2\pi -\frac{1}{2}\log |U_{k}| -\frac{1}{2}\log |V| - \frac{1}{2}\hat{b}_{i}^{T}V^{-1}\hat{b}_{i} + \hat{b}_{i}^{T}V^{-1}\mathbb{E}(b_{i}|\hat{b}_{i}) -\frac{1}{2}tr\left(V^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right) - \frac{1}{2}tr\left(U_{k}^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right)
\end{align*}
$$

V has a specific form:
$$
V = \left( \begin{matrix}1 & \rho \\ \rho & 1 \end{matrix} \right)
$$
Let $\mu_{i} = \mathbb{E}(b_{i}|\hat{b}_{i})$
$$
\begin{align*}
\mathbb{E}_{b_{i}|\hat{b}_{i}}\left[ \log N(\hat{b}_{i}; b_{i}, V) + \log N(b_{i}; 0, U_{k}) \right] &= -2\log 2\pi -\frac{1}{2}\log |U_{k}| -\frac{1}{2}\log(1-\rho^2) - \frac{1}{2(1-\rho^2)}\left(\hat{b}_{i1}^2 + \hat{b}_{i2}^2 -2\hat{b}_{i1} \mu_{i1} -2\hat{b}_{i2} \mu_{i2} + \mathbb{E}(b_{i1}^2|\hat{b}_{i}) + \mathbb{E}(b_{i2}^2|\hat{b}_{i}) - 2\hat{b}_{i1}\hat{b}_{i2}\rho + 2 \hat{b}_{i1}\mu_{i2}\rho +2\hat{b}_{i2}\mu_{i1}\rho - 2\rho\mathbb{E}(b_{i1}b_{i2}|\hat{b}_{i}) \right) -  \frac{1}{2}tr\left(U_{k}^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right)
\end{align*}
$$

$$
\gamma_{z_{i}}(k) = P(z_{i}=k|X_{i}) = \frac{\pi_{k}N(x_{i}; 0, V+U_{k})}{\sum_{k'=0}^{K}\pi_{k'}N(x_{i}; 0, V + U_{k'})}
$$

### M step

$V$:
$$
\begin{align*}
f(V^{-1}) = \sum_{i=1}^{n} \sum_{k=0}^{K} \gamma_{Z_{i}}(k)\left[ -p\log 2\pi -\frac{1}{2}\log |U_{k}| -\frac{1}{2}\log |V| - \frac{1}{2}\hat{b}_{i}^{T}V^{-1}\hat{b}_{i} + \hat{b}_{i}^{T}V^{-1}\mathbb{E}(b_{i}|\hat{b}_{i}) -\frac{1}{2}tr\left(V^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right) - \frac{1}{2}tr\left(U_{k}^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right) \right]
\end{align*}
$$

$$
\begin{align*}
f(V^{-1})' &= \sum_{i=1}^{n} \sum_{k=0}^{K} \gamma_{Z_{i}}(k)\left[ \frac{1}{2}V - \frac{1}{2}\hat{b}_{i}\hat{b}_{i}^{T} + \mathbb{E}(b_{i}|\hat{b}_{i})\hat{b}_{i}^{T} - \frac{1}{2} \mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right] = 0 \\
\frac{1}{2}Vn &= \sum_{i=1}^{n} \sum_{k=0}^{K} \gamma_{Z_{i}}(k)\left[\frac{1}{2}\hat{b}_{i}\hat{b}_{i}^{T} - \mathbb{E}(b_{i}|\hat{b}_{i})\hat{b}_{i}^{T} + \frac{1}{2} \mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right] \\
\hat{V} &= \frac{1}{n} \sum_{i=1}^{n} \left[\hat{b}_{i}\hat{b}_{i}^{T} - 2\mathbb{E}(b_{i}|\hat{b}_{i})\hat{b}_{i}^{T} + \mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right] \\
&= \frac{1}{n} \sum_{i=1}^{n} \mathbb{E}\left[ (\hat{b}_{i} - b_{i})(\hat{b}_{i} - b_{i})^{T} | \hat{b}_{i}\right]
\end{align*}
$$


$\rho$:
$$
f(\rho) = \sum_{i=1}^{n} \sum_{k=1}^{K} \gamma_{Z_{i}}(k)\left[-2\log 2\pi -\frac{1}{2}\log |U_{k}| -\frac{1}{2}\log(1-\rho^2) - \frac{1}{2(1-\rho^2)}\left(\hat{b}_{i1}^2 + \hat{b}_{i2}^2 -2\hat{b}_{i1} \mu_{i1} -2\hat{b}_{i2} \mu_{i2} + \mathbb{E}(b_{i1}^2|\hat{b}_{i}) + \mathbb{E}(b_{i2}^2|\hat{b}_{i}) - 2\hat{b}_{i1}\hat{b}_{i2}\rho + 2 \hat{b}_{i1}\mu_{i2}\rho +2\hat{b}_{i2}\mu_{i1}\rho - 2\rho\mathbb{E}(b_{i1}b_{i2}|\hat{b}_{i}) \right) -  \frac{1}{2}tr\left(U_{k}^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i}) \right)\right]
$$

$$
\begin{align*}
f(\rho)' = \sum_{i=1}^{n} \sum_{k=1}^{K} \gamma_{Z_{i}}(k)\left[ \frac{\rho}{1-\rho^2} -\frac{\rho}{(1-\rho^2)^2}\left( \hat{b}_{i1}^2 + \hat{b}_{i2}^2 -2\hat{b}_{i1} \mu_{i1} -2\hat{b}_{i2} \mu_{i2} + \mathbb{E}(b_{i1}^2|\hat{b}_{i}) + \mathbb{E}(b_{i2}^2|\hat{b}_{i}) \right) -\frac{\rho^2+1}{(1-\rho^2)^2}\left( -\hat{b}_{i1}\hat{b}_{i2} + \hat{b}_{i1}\mu_{i2} +\hat{b}_{i2}\mu_{i1} - \mathbb{E}(b_{i1}b_{i2}|\hat{b}_{i}) \right) \right] &= 0 \\
\rho(1-\rho^2)n - \rho \sum_{i=1}^{n} \sum_{k=1}^{K} \gamma_{Z_{i}}(k) \left( \hat{b}_{i1}^2 + \hat{b}_{i2}^2 -2\hat{b}_{i1} \mu_{i1} -2\hat{b}_{i2} \mu_{i2} + \mathbb{E}(b_{i1}^2|\hat{b}_{i}) + \mathbb{E}(b_{i2}^2|\hat{b}_{i}) \right) - (\rho^2 + 1) \sum_{i=1}^{n} \sum_{k=1}^{K} \gamma_{Z_{i}}(k)\left( -\hat{b}_{i1}\hat{b}_{i2} + \hat{b}_{i1}\mu_{i2} +\hat{b}_{i2}\mu_{i1} - \mathbb{E}(b_{i1}b_{i2}|\hat{b}_{i}) \right) &= 0 \\
-n\rho^{3} - \rho^2 \sum_{i=1}^{n} \left( -\hat{b}_{i1}\hat{b}_{i2} + \hat{b}_{i1}\mu_{i2} +\hat{b}_{i2}\mu_{i1} - \mathbb{E}(b_{i1}b_{i2}|\hat{b}_{i}) \right) - \rho \sum_{i=1}^{n} \left( \hat{b}_{i1}^2 + \hat{b}_{i2}^2 -2\hat{b}_{i1} \mu_{i1} -2\hat{b}_{i2} \mu_{i2} + \mathbb{E}(b_{i1}^2|\hat{b}_{i}) + \mathbb{E}(b_{i2}^2|\hat{b}_{i}) -1\right) - \sum_{i=1}^{n} \left( -\hat{b}_{i1}\hat{b}_{i2} + \hat{b}_{i1}\mu_{i2} +\hat{b}_{i2}\mu_{i1} - \mathbb{E}(b_{i1}b_{i2}|\hat{b}_{i}) \right) &= 0
\end{align*}
$$

The polynomial has either 1 or 3 real roots in (-1, 1).

Algorithm: 
```{r, eval=FALSE, tidy=FALSE, highlight=FALSE}
Input: X, Ulist, init_rho
Compute loglikelihood
delta = 1
while delta > tol
  Given rho, Estimate pi using convex method (current mash method)
  M step: update rho: find all roots of polynomial, if it has three real roots, choose the one with higher loglikelihood.
  Compute loglikelihood
  Update delta
```

```{r}
mixture.EM.times <- function(X, Ulist, init_rho=0, tol=1e-5, prior = c('nullbiased', 'uniform'), grid = 1,...){
  times = length(init_rho)
  result = list()
  loglik = c()
  rho = c()
  time.t = c()
  converge.status = c()
  for(i in 1:times){
    if(is.null(grid)){
      out.time = system.time(result[[i]] <- mixture.EM(X, Ulist,
                                                        init_rho=init_rho[i],
                                                        prior=prior,
                                                        tol = tol))
    }else{
      out.time = system.time(result[[i]] <- mixture.EM(X, Ulist,
                                                        init_rho=init_rho[i],
                                                        prior=prior,
                                                        tol = tol,
                                                        grid = grid, ...))
    }

    time.t = c(time.t, out.time['elapsed'])
    loglik = c(loglik, tail(result[[i]]$log_liks, 1))
    rho = c(rho, result[[i]]$rho)
  }
  if(abs(max(loglik) - min(loglik)) < 1e-4){
    status = 'global'
  }else{
    status = 'local'
  }
  ind = which.max(loglik)
  return(list(result = result[[ind]], status = status, loglik = loglik, rho=rho, time = time.t))
}

mixture.EM <- function(X, Ulist, init_rho=0, tol=1e-5, prior = c('nullbiased', 'uniform'), grid=1, usepointmass=FALSE,...){
  prior <- match.arg(prior)
  prior.v <- mashr:::set_prior(length(Ulist), prior)
  # estimate pi
  m.model = fit_mash(X, Ulist, rho = init_rho, prior=prior, grid=grid,usepointmass=usepointmass, ...)
  pi_s = get_estimated_pi(m.model)
  # compute loglikelihood
  log_liks <- c()
  log_liks <- c(log_liks, get_loglik(m.model)+penalty(prior.v, pi_s))
  delta.ll <- 1
  niter <- 0
  while(delta.ll > tol){
    # max_rho
    rho_s <- E_rho(X, m.model)
    if(length(rho_s) == 1){
      m.model = fit_mash(X, Ulist, rho_s, prior=prior, grid=grid, usepointmass=usepointmass,...)
      pi_s = get_estimated_pi(m.model)
      log_liks <- c(log_liks, get_loglik(m.model)+penalty(prior.v, pi_s))
      # Update delta
      delta.ll <- log_liks[length(log_liks)] - log_liks[length(log_liks)-1]
      niter <- niter + 1
    }else{
      temp.ll = c()
      temp.model = list()
      for(i in 1:3){
        temp.m = fit_mash(X, Ulist, rho_s[i], prior = prior, grid=grid, ...)
        temp.ll = c(temp.ll, get_loglik(temp.m)+penalty(prior.v, pi_s))
        temp.model = c(temp.model, list(temp.m))
      }
      rho_s = rho_s[which.max(temp.ll)]
      m.model = temp.model[[which.max(temp.ll)]]
      pi_s = get_estimated_pi(m.model)
      log_liks <- c(log_liks, get_loglik(m.model)+penalty(prior.v, pi_s))
      # Update delta
      delta.ll <- log_liks[length(log_liks)] - log_liks[length(log_liks)-1]
      niter <- niter + 1
    }
  }
  return(list(pi = pi_s, rho = rho_s, log_liks = log_liks))
}

E_rho <- function(X, m.model){
  n = nrow(X)
  post.m = m.model$result$PosteriorMean
  post.sec = array(unlist(lapply(1:n, function(i) m.model$result$PosteriorCov[,,i] + post.m[i,] %*% t(post.m[i,]))),
                   dim=c(2, 2, n))
  temp2 = -sum(X[,1]*X[,2]) + sum(X[,1]*post.m[,2]) + sum(X[,2]*post.m[,1]) - sum(post.sec[1,2,])
  temp1 = sum(X[,1]^2 + X[,2]^2) - 2*sum(X[,1]*post.m[,1]) - 2*sum(X[,2]*post.m[,2]) + sum(post.sec[1,1,] + post.sec[2,2,])

  rts = polyroot(c(temp2, temp1-n, temp2, n))
  
  # check complex number
  is.real = abs(Im(rts))<1e-12
  if(sum(is.real) == 1){
    return(Re(rts[is.real]))
  }else{
    return(Re(rts))
  }
}

fit_mash <- function(X, Ulist, rho, prior=c('nullbiased', 'uniform'), grid = 1, ...){
  m.data = mashr::mash_set_data(Bhat=X, V = matrix(c(1, rho, rho, 1), 2, 2))
  m.model = mashr::mash(m.data, Ulist, prior=prior, verbose = FALSE, outputlevel = 3, grid = grid, ...)
  return(m.model)
}
```

## Data

$$
\hat{\beta}|\beta \sim N_{2}(\hat{\beta}; \beta, \left(\begin{matrix} 1 & 0.5 \\
                                          0.5 & 1 \end{matrix}\right))
$$

$$
\beta \sim \frac{1}{4}\delta_{0} + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 0 \\
                                          0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 0 & 0 \\
                                          0 & 1 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 1 \\
                                          1 & 1 \end{matrix}\right))
$$

n = 4000

```{r}
set.seed(1)
n = 4000; p = 2
Sigma = matrix(c(1,0.5,0.5,1),p,p)
U0 = matrix(0,2,2)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[2,2] = 1
U3 = matrix(1,2,2)
Utrue = list(U0=U0, U1=U1, U2=U2, U3=U3)
data = generate_data(n, p, Sigma, Utrue)
```

```{r}
m.data = mash_set_data(data$Bhat, data$Shat)
U.c = cov_canonical(m.data)
grid = mashr:::autoselect_grid(m.data, sqrt(2))
Ulist = mashr:::normalize_Ulist(U.c)
xUlist = mashr:::expand_cov(Ulist,grid,usepointmass =  TRUE)
result <- mixture.EM2.times(data$Bhat, xUlist, init_rho = c(-0.7,0,0.7), grid=1)
plot(result$result$log_liks)
```
The estimated $\rho$ is `r result[[1]]$rho`.

```{r}
m.data.em = mash_set_data(data$Bhat, data$Shat, V = matrix(c(1,result[[1]]$rho,result[[1]]$rho,1),2,2))
U.c = cov_canonical(m.data.em)
m.em = mash(m.data.em, U.c, verbose= FALSE)
null.ind = which(apply(data$B,1,sum) == 0)
```

The log likelihood is `r round(get_loglik(m.em),2)`. There are `r length(get_significant_results(m.em))` significant samples, `r sum(get_significant_results(m.em) %in% null.ind)` false positives. The RRMSE is `r RRMSE(data$B, data$Bhat, list(m.em))`.

The estimated `pi` is
```{r}
barplot(get_estimated_pi(m.em), las=2, cex.names = 0.7, main='EM rho', ylim=c(0,0.8))
```

The ROC curve:
```{r}
m.data.correct = mash_set_data(data$Bhat, data$Shat, V=Sigma)
m.correct = mash(m.data.correct, U.c, verbose = FALSE)
m.correct.seq = ROC.table(data$B, m.correct)
m.em.seq = ROC.table(data$B, m.em)
```

```{r echo=FALSE, fig.align = "center"}
{plot(m.correct.seq[,'FPR'], m.correct.seq[,'TPR'],type='l',xlab = 'FPR', ylab='TPR', main='True Positive vs False Positive', cex=1.5, lwd = 1.5)
lines(m.em.seq[,'FPR'], m.em.seq[,'TPR'], col='red', lwd = 1.5)
legend('bottomright', c('True','EM'),col=c('black','red'),lty=c(1,1), lwd=c(1.5,1.5))}
```

# Another EM version for rho optim

Algorithm: 
```{r, eval=FALSE, tindy = FALSE}
Input: X, Ulist, init_rho
Given rho, estimate pi by max loglikelihood (convex problem)
Compute loglikelihood
delta = 1
while delta > tol
  E step: compute z
  M step: update rho
  Given rho, estimate pi by max loglikelihood (convex problem)
  Compute loglikelihood
  Update delta
```

```{r}
mixture.M.times <- function(X, Ulist, init_rho=0, init_pi=NULL, prior = c('nullbiased', 'uniform'), control = list()){
  times = length(init_rho)
  result = list()
  loglik = c()
  rho = c()
  time.t = c()
  converge.status = c()
  for(i in 1:times){
    out.time = system.time(result[[i]] <- mixture.M(X, Ulist,
                                                     init_pi=init_pi,
                                                     init_rho=init_rho[i],
                                                     prior=prior, 
                                                     control = control))
    time.t = c(time.t, out.time['elapsed'])
    loglik = c(loglik, -result[[i]]$B)
    rho = c(rho, result[[i]]$rhohat)
    converge.status = c(converge.status, result[[i]]$converged)
  }
  if(abs(max(loglik) - min(loglik)) < 1e-4){
    status = 'global'
  }else{
    status = 'local'
  }
  ind = which.max(loglik)
  return(list(result = result[[ind]], status = status, loglik = loglik, rho=rho, time = time.t, converge.status = converge.status))
}

mixture.M <- function(X, Ulist, init_rho=0, tol=1e-5, prior = c('nullbiased', 'uniform')) {
  prior <- match.arg(prior)

  m.model = fit_mash(X, Ulist, rho = init_rho, prior=prior)
  pi_s = get_estimated_pi(m.model, dimension = 'all')
  # get complete Ulist
  xUlist = mashr:::expand_cov(Ulist, m.model$fitted_g$grid, m.model$fitted_g$usepointmass)
  prior.v <- mashr:::set_prior(length(pi_s), prior)
  
  # compute loglikelihood
  log_liks <- c()
  log_liks <- c(log_liks, get_loglik(m.model)+penalty(prior.v, pi_s))
  delta.ll <- 1
  niter <- 0
  rho = init_rho
  
  while(delta.ll > tol){
    # compute L
    Sigma <- get_sigma(rho, xUlist)
    L <- t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U)}))
  
    m  = t(pi_s * t(L)) # matrix_lik is n by k; so this is also n by k
    m.rowsum = rowSums(m)
    classprob = m/m.rowsum #an n by k matrix
  
    # max_rho
    rho <- optimize(EMloglikelihood, interval = c(-1,1), maximum = TRUE, X = X, Ulist = xUlist, z = classprob)$maximum
    
    m.model = fit_mash(X, Ulist, rho, prior=prior)
    
    pi_s = get_estimated_pi(m.model, dimension = 'all')
    log_liks <- c(log_liks, get_loglik(m.model)+penalty(prior.v, pi_s))
    # Update delta
    delta.ll <- log_liks[length(log_liks)] - log_liks[length(log_liks)-1]
    niter <- niter + 1
  }

  return(list(pihat = normalize(pi_s), rho = rho, loglik=log_liks))
}

EMloglikelihood = function(rho, X, Ulist, z){
  Sigma = get_sigma(rho, Ulist)
  L = t(plyr::laply(Sigma,function(U){mvtnorm::dmvnorm(x=X,sigma=U, log=TRUE)}))
  sum(L * z)
}

fit_mash <- function(X, Ulist, rho, prior=c('nullbiased', 'uniform')){
  m.data = mashr::mash_set_data(Bhat=X, Shat=1, V = matrix(c(1, rho, rho, 1), 2, 2))
  m.model = mashr::mash(m.data, Ulist, prior=prior, verbose = FALSE)
  return(m.model)
}

```

# Data

```{r}
set.seed(1)
n = 4000; p = 2
Sigma = matrix(c(1,0.5,0.5,1),p,p)
U0 = matrix(0,2,2)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[2,2] = 1
U3 = matrix(1,2,2)
Utrue = list(U0=U0, U1=U1, U2=U2, U3=U3)
data = generate_data(n, p, Sigma, Utrue)
```

```{r}
m.data = mash_set_data(data$Bhat, data$Shat)
U.c = cov_canonical(m.data)

result.m <- mixture.M(m.data$Bhat, U.c)
```

The estimated $\rho$ is `r result.m$rho`.

```{r}
m.data.m = mash_set_data(data$Bhat, data$Shat, V = matrix(c(1,result.m$rho,result.m$rho,1),2,2))
U.c.m = cov_canonical(m.data.m)
m.m = mash(m.data.m, U.c, verbose= FALSE)
null.ind = which(apply(data$B,1,sum) == 0)
```

The log likelihood is `r round(get_loglik(m.m),2)`. There are `r length(get_significant_results(m.m))` significant samples, `r sum(get_significant_results(m.m) %in% null.ind)` false positives. The RRMSE is `r RRMSE(data$B, data$Bhat, list(m.m))`.
