---
title: "Estimate Null Correlation Problem"
author: "Yuxin Zou"
date: 2018-07-09
output: 
  html_document: 
    code_folding: hide
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

```{r library}
library(mashr)
```

In my simple simulation, the current approach underestimate the null correlation. We need to find better positive definite estimator. We could try to estimate the pairwise correlation, ie. mle of $\sum{\pi_{lk} N_{2}(0, V + w_{l}U_{k})}$ for any pair of conditions.

Simple simulation in $R^2$ to illustrate the problem:
$$
\hat{\beta}|\beta \sim N_{2}(\hat{\beta}; \beta, \left(\begin{matrix} 1 & 0.5 \\
                                          0.5 & 1 \end{matrix}\right))
$$

$$
\beta \sim \frac{1}{4}\delta_{0} + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 0 \\
                                          0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 0 & 0 \\
                                          0 & 1 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 1 \\
                                          1 & 1 \end{matrix}\right))
$$

$\Rightarrow$
$$
\hat{\beta} \sim \frac{1}{4}N_{2}(0, \left( \begin{matrix} 1 & 0.5 \\
0.5 & 1 \end{matrix} \right)) + \frac{1}{4}N_{2}(0, \left( \begin{matrix} 2 & 0.5 \\
0.5 & 1 \end{matrix} \right)) + \frac{1}{4}N_{2}(0, \left( \begin{matrix} 1 & 0.5 \\
0.5 & 2 \end{matrix} \right)) + \frac{1}{4}N_{2}(0, \left( \begin{matrix} 2 & 1.5 \\
1.5 & 2 \end{matrix} \right))
$$

```{r}
set.seed(1)
n = 500; p = 2
Sigma = matrix(c(1,0.5,0.5,1),p,p)
beta.0 = beta.1 = beta.2 = beta.3 = matrix(0,n,p)
beta.1[,1] = rnorm(n)
beta.2[,2] = rnorm(n)
beta.3[,1] = rnorm(n)
beta.3[,2] = beta.3[,1]
Beta = rbind(beta.0, beta.1, beta.2, beta.3)
E = MASS::mvrnorm(n*4, rep(0, p), Sigma)
Bhat = Beta + E
SE = 1
```

Using truncated empirical correlation 
```{r}
mash.data = mash_set_data(Bhat, SE)
Vhat = estimate_null_correlation(mash.data, apply_lower_bound = FALSE)
Vhat
```

We underestimate the correlation among measurement errors.

Let's see the result from mash

```{r}
# Use underestimate cor
mash.data.V = mash_set_data(Bhat, SE, V=Vhat)
U.c = cov_canonical(mash.data.V)
m.V = mash(mash.data.V, U.c)
barplot(get_estimated_pi(m.V), las=2, cex.names = 0.7)
```
The loglikelihood is
```{r}
get_loglik(m.V)
```

Let's see what will happen if we overestimate correlation
```{r}
# If we overestimate cor
V.o = matrix(c(1,0.75,0.75,1),2,2)
mash.data.Vo = mash_set_data(Bhat, SE, V=V.o)
m.Vo = mash(mash.data.Vo, U.c)
barplot(get_estimated_pi(m.Vo), las=2, cex.names = 0.7)
```
The loglikelihood is 
```{r}
get_loglik(m.Vo)
```

What will happen if we have correct correlation
```{r}
# With correct cor
mash.data.correct = mash_set_data(Bhat, SE, V=Sigma)
m.correct = mash(mash.data.correct, U.c)
barplot(get_estimated_pi(m.correct), las=2, cex.names = 0.7)
```

The loglikelihood is 
```{r}
get_loglik(m.correct)
```

# Sol1: mash 1 by 1

We run ash for each condition, and estimate correlation matrix based on the non-significant genes. The estimated cor is closer to the truth. It overestimates the cor.
```{r}
m.1by1 = mash_1by1(mash.data)
strong = get_significant_results(m.1by1)
V.mash = cor(Bhat[-strong,])
V.mash
```

```{r}
mash.data.1by1 = mash_set_data(Bhat, SE, V=V.mash)
m.V1by1 = mash(mash.data.1by1, U.c)
barplot(get_estimated_pi(m.V1by1), las=2, cex.names = 0.7)
```

The loglikelihood is
```{r}
get_loglik(m.V1by1)
```

# Sol2: MLE

## K=1

Suppose a simple extreme case
$$
\left(\begin{matrix} \hat{x} \\ \hat{y} \end{matrix} \right)| \left(\begin{matrix} x \\ y \end{matrix} \right) \sim N_{2}(\left(\begin{matrix} \hat{x} \\ \hat{y} \end{matrix} \right); \left(\begin{matrix} x \\ y \end{matrix} \right), \left( \begin{matrix} 1 & \rho \\ \rho & 1 \end{matrix}\right))
$$
$$
\left(\begin{matrix} x \\ y \end{matrix} \right) \sim \delta_{0}
$$
$\Rightarrow$
$$
\left(\begin{matrix} \hat{x} \\ \hat{y} \end{matrix} \right) \sim N_{2}(\left(\begin{matrix} \hat{x} \\ \hat{y} \end{matrix} \right); \left(\begin{matrix} 0 \\ 0 \end{matrix} \right), \left( \begin{matrix} 1 & \rho \\ \rho & 1 \end{matrix}\right))
$$

$$
f(\hat{x},\hat{y}) = \prod_{i=1}^{n} \frac{1}{2\pi\sqrt{1-\rho^2}} \exp \{-\frac{1}{2(1-\rho^2)}\left[ \hat{x}_{i}^2 + \hat{y}_{i}^2 - 2\rho \hat{x}_{i}\hat{y}_{i}\right]  \}
$$
The MLE of $\rho$:
$$
\begin{align*}
l(\rho) &= -\frac{n}{2}\log(1-\rho^2) - \frac{1}{2(1-\rho^2)}\left( \sum_{i=1}^{n} x_{i}^2 + y_{i}^2 - 2\rho x_{i}y_{i} \right) \\
l(\rho)' &= \frac{n\rho}{1-\rho^2} - \frac{\rho}{(1-\rho^2)^2} \sum_{i=1}^{n} (x_{i}^2 + y_{i}^2) + \frac{\rho^2 + 1}{(1-\rho^2)^2} \sum_{i=1}^{n} x_{i}y_{i} = 0 \\
&= \rho^{3} - \rho^{2}\frac{1}{n}\sum_{i=1}^{n} x_{i}y_{i} - \left( 1- \frac{1}{n} \sum_{i=1}^{n} x_{i}^{2} + y_{i}^{2} \right) \rho - \frac{1}{n}\sum_{i=1}^{n} x_{i}y_{i} = 0 \\
l(\rho)'' &= \frac{n(\rho^2+1)}{(1-\rho^2)^2} - \frac{1}{2}\left( \frac{8\rho^2}{(1-\rho^2)^{3}} + \frac{2}{(1-\rho^2)^2} \right)\sum_{i=1}^{n}(x_{i}^2 + y_{i}^2) + \{ \left( \frac{8\rho^2}{(1-\rho^2)^{3}} + \frac{2}{(1-\rho^2)^2} \right)\rho + \frac{4\rho}{(1-\rho^2)^2} \}\sum_{i=1}^{n}x_{i}y_{i}
\end{align*}
$$

**The log likelihood is not a concave function in general.** The score function has either 1 or 3 real solutions. 

Kendall and Stuart (1979) noted that at least one of the roots is real and lies in
the interval [−1, 1]. However, it is possible that all three roots are real and in the admissible interval, in which case the likelihood can be evaluated at each root to determine the true maximum likelihood estimate. 

I simulate the data with $\rho=0.75$ and plot the loglikelihood function:

```{r echo=FALSE}
n = 1000; rho = 0.75
V = matrix(c(1,rho,rho,1),2,2)
U = matrix(0,2,2)
set.seed(1)
X.0 = MASS::mvrnorm(n, rep(0, 2), Sigma = V+U)
z.null = X.0

rho.seq = seq(-0.99,0.99,by=0.01)
loglik = numeric(199)
for(i in 1:199){
  V = matrix(c(1,rho.seq[i], rho.seq[i], 1), 2, 2)
  loglik[i] = sum(mvtnorm::dmvnorm(x=z.null, sigma=V+U, log=TRUE))
}
plot(rho.seq, loglik, type='l', ylab = 'log likelihood')
```

$l(\rho)'$ has one real solution
```{r}
polyroot(c(- sum(z.null[,1]*z.null[,2]),  - (n - sum(z.null[,1]^2 + z.null[,2]^2)), - sum(z.null[,1]*z.null[,2]), n))
```

## In general

The general derivation is in [estimate correlation mle](EstimateCorMax.html)

# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
