---
title: "Estimate cor — M V decompose"
author: "Yuxin Zou"
date: 2018-10-5
output: 
  workflowr::wflow_html:
    code_folding: hide
---

```{r}
library(mashr)
source('../code/generateDataV.R')
source('../code/summary.R')
```

We use EM algorithm to update $\rho$.

B is the $n\times R$ true value matrix. $\mathbf{z}$ is a length n vector.

## E step

$$
P(\hat{B},B|\rho, \pi) = \prod_{i=1}^{n} \left[N(\hat{b}_{i}; b_{i}, S_{i}VS_{i})\sum_{p=0}^{P} \pi_{p} N(b_{i}; 0, \Sigma_{p})\right]
$$

$$
\begin{align*}
\mathbb{E}_{B|\hat{B}} \log P(\hat{B},B|\rho, \pi) &= \sum_{i=1}^{n} \mathbb{E}_{b_{i}|\hat{b}_{i}}\left[ \log N(\hat{b}_{i}; b_{i}, S_{i}VS_{i}) + \log \sum_{p=0}^{P} \pi_{p} N(b_{i}; 0, \Sigma_{p}) \right] \\
&= \sum_{i=1}^{n} \mathbb{E}_{b_{i}|\hat{b}_{i}}\log N(\hat{b}_{i}; b_{i}, S_{i}VS_{i}) + \sum_{i=1}^{n}\mathbb{E}_{b_{i}|\hat{b}_{i}}\log \sum_{p=0}^{P} \pi_{p} N(b_{i}; 0, \Sigma_{p})
\end{align*}
$$

$V$ depends on the first term only. Let $\mu_{i} = \mathbb{E}_{b_{i}|\hat{b}_{i}}(b_{i})$
$$
\begin{align*}
\log N(\hat{b}_{i}; b_{i}, V) &= -\frac{R}{2}\log 2\pi -\frac{1}{2}\log |S_{i}VS_{i}| - \frac{1}{2}(\hat{b}_{i}-b_{i})^{T}S_{i}^{-1}V^{-1}S_{i}^{-1}(\hat{b}_{i}-b_{i}) \\
&= -\frac{R}{2}\log 2\pi -\frac{1}{2}\log |S_{i}VS_{i}| - \frac{1}{2}\hat{b}_{i}^{T}S_{i}^{-1}V^{-1}S_{i}^{-1}\hat{b}_{i} + \frac{1}{2}b_{i}^{T}S_{i}^{-1}V^{-1}S_{i}^{-1}\hat{b}_{i} + \frac{1}{2}\hat{b}_{i}^{T}S_{i}^{-1}V^{-1}S_{i}^{-1}b_{i} -\frac{1}{2}b_{i}^{T}S_{i}^{-1}V^{-1}S_{i}^{-1}b_{i} \\
\mathbb{E}_{b_{i}|\hat{b}_{i}} \log N(\hat{b}_{i}; b_{i}, V) &= -\frac{R}{2}\log 2\pi -\frac{1}{2}\log |S_{i}VS_{i}| - \frac{1}{2}\hat{b}_{i}^{T}S_{i}^{-1}V^{-1}S_{i}^{-1}\hat{b}_{i} + \frac{1}{2}\mu_{i}^{T}S_{i}^{-1}V^{-1}S_{i}^{-1}\hat{b}_{i} + \frac{1}{2}\hat{b}_{i}^{T}S_{i}^{-1}V^{-1}S_{i}^{-1}\mu_{i} -\frac{1}{2}tr(S_{i}^{-1}V^{-1}S_{i}^{-1}\mathbb{E}_{b_{i}|\hat{b}_{i}}(b_{i}b_{i}^{T}))
\end{align*}
$$

Maximize with respect to V: 

We have constraint on V, the diagonal of V must be 1. Let $V = DCD$, C is the covariance matrix, D = $diag(1/sqrt(C_{jj}))$.

$$
f(C) = \sum_{i=1}^{n} -\frac{R}{2}\log 2\pi -\frac{1}{2}\log |C|- \log |DS_{i}| - \frac{1}{2}\hat{b}_{i}^{T}S_{i}^{-1}D^{-1}C^{-1}D^{-1}S_{i}^{-1}\hat{b}_{i} + \frac{1}{2}\mu_{i}^{T}S_{i}^{-1}D^{-1}C^{-1}D^{-1}S_{i}^{-1}\hat{b}_{i} + \frac{1}{2}\hat{b}_{i}^{T}S_{i}^{-1}D^{-1}C^{-1}D^{-1}S_{i}^{-1}\mu_{i} -\frac{1}{2}tr(S_{i}^{-1}D^{-1}C^{-1}D^{-1}S_{i}^{-1}\mathbb{E}_{b_{i}|\hat{b}_{i}}(b_{i}b_{i}^{T}))
$$

$$
\begin{align*}
f(C)' &= \sum_{i=1}^{n} -\frac{1}{2}C^{-1} + \frac{1}{2}C^{-1}D^{-1}S_{i}^{-1}\hat{b}_{i}\hat{b}_{i}^{T}S_{i}^{-1}D^{-1}C^{-1} - \frac{1}{2} C^{-1}D^{-1}S_{i}^{-1}\mu_{i}\hat{b}_{i}^{T}S_{i}^{-1}D^{-1}C^{-1} - \frac{1}{2}C^{-1}D^{-1}S_{i}^{-1}\hat{b}_{i}\mu_{i}^{T}S_{i}^{-1}D^{-1}C^{-1} + \frac{1}{2} C^{-1}D^{-1}S_{i}^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i})S_{i}^{-1}D^{-1}C^{-1} = 0 \\
0 &= \sum_{i=1}^{n} -\frac{1}{2}C + \frac{1}{2}D^{-1}S_{i}^{-1}\hat{b}_{i}\hat{b}_{i}^{T}S_{i}^{-1}D^{-1} - \frac{1}{2}D^{-1}S_{i}^{-1}\mu_{i}\hat{b}_{i}^{T}S_{i}^{-1}D^{-1} - \frac{1}{2}D^{-1}S_{i}^{-1}\hat{b}_{i}\mu_{i}^{T}S_{i}^{-1}D^{-1} + \frac{1}{2} D^{-1}S_{i}^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i})S_{i}^{-1}D^{-1} \\
\hat{C} &= \frac{1}{n} \sum_{i=1}^{n} \left[D^{-1}S_{i}^{-1}\hat{b}_{i}\hat{b}_{i}^{T}S_{i}^{-1}D^{-1} - D^{-1}S_{i}^{-1}\mu_{i}\hat{b}_{i}^{T}S_{i}^{-1}D^{-1} - D^{-1}S_{i}^{-1}\hat{b}_{i}\mu_{i}^{T}S_{i}^{-1}D^{-1} + D^{-1}S_{i}^{-1}\mathbb{E}(b_{i}b_{i}^{T}|\hat{b}_{i})S_{i}^{-1}D^{-1} \right] \\
&= \frac{1}{n} \sum_{i=1}^{n} \mathbb{E}\left[ (D^{-1}S_{i}^{-1}(\hat{b}_{i} - b_{i}))(D^{-1}S_{i}^{-1}(\hat{b}_{i} - b_{i}))^{T} | \hat{b}_{i}\right] \\
&= D^{-1}\frac{1}{n} \sum_{i=1}^{n} S_{i}^{-1}\mathbb{E}\left[ (\hat{b}_{i} - b_{i})(\hat{b}_{i} - b_{i})^{T} | \hat{b}_{i}\right] S_{i}^{-1} D^{-1}
\end{align*}
$$

We can update C and V as
$$
\hat{C}_{(t+1)} =  \hat{D}^{-1}_{(t)}\frac{1}{n} \left[\sum_{i=1}^{n} S_{i}^{-1}\mathbb{E}\left[ (\hat{b}_{i} - b_{i})(\hat{b}_{i} - b_{i})^{T} | \hat{b}_{i}\right]S_{i}^{-1} \right] \hat{D}^{-1}_{(t)} \\
\hat{D}_{(t+1)} = diag(1/\sqrt{\hat{C}_{(t+1)jj}}) \\
\hat{V}_{(t+1)} = \hat{D}_{(t+1)}\hat{C}_{(t+1)}\hat{D}_{(t+1)}
$$

The resulting $\hat{V}_{(t+1)}$ is equivalent as
$$
\hat{C}_{(t+1)} = \frac{1}{n}\sum_{i=1}^{n} S_{i}^{-1}\mathbb{E}\left[ (\hat{b}_{i} - b_{i})(\hat{b}_{i} - b_{i})^{T} | \hat{b}_{i}\right]S_{i}^{-1} \\
\hat{D}_{(t+1)} = diag(1/\sqrt{\hat{C}_{(t+1)jj}}) \\
\hat{V}_{(t+1)} = \hat{D}_{(t+1)}\hat{C}_{(t+1)}\hat{D}_{(t+1)}
$$

It is hard to estimate $\boldsymbol{\pi}$ from the second term, $\sum_{i=1}^{n}\mathbb{E}_{b_{i}|\hat{b}_{i}}\log \sum_{p=0}^{P} \pi_{p} N(b_{i}; 0, \Sigma_{p})$.

Given V, we estimate $\boldsymbol{\pi}$ by max loglikelihood, which is a convex problem

Algorithm: 
```{r, eval=FALSE, tindy = FALSE}
Input: X, Ulist, init_V
Given V, estimate pi by max loglikelihood (convex problem)
Compute loglikelihood
delta = 1
while delta > tol
  M step: update C
  Convert to V
  Given V, estimate pi by max loglikelihood (convex problem)
  Compute loglikelihood
  Update delta
```

```{r}
penalty <- function(prior, pi_s){
  subset <- (prior != 1.0)
  sum((prior-1)[subset]*log(pi_s[subset]))
}

mixture.MV <- function(mash.data, Ulist, init_V=diag(ncol(mash.data$Bhat)), max_iter = 500, tol=1e-5, prior = c('nullbiased', 'uniform'), cor = TRUE, track_fit = FALSE){
  prior <- match.arg(prior)
  tracking = list()

  m.model = fit_mash_V(mash.data, Ulist, V = init_V, prior=prior)
  pi_s = get_estimated_pi(m.model, dimension = 'all')
  prior.v <- mashr:::set_prior(length(pi_s), prior)
  
  # compute loglikelihood
  log_liks <- numeric(max_iter+1)
  log_liks[1] <- get_loglik(m.model)+penalty(prior.v, pi_s)
  V = init_V
  
  result = list(V = V, logliks = log_liks[1], mash.model = m.model)
  
  for(i in 1:max_iter){
    if(track_fit){
      tracking[[i]] = result
    }
    # max_V
    V = E_V(mash.data, m.model)
    if(cor){
        V = cov2cor(V)
    }
    m.model = fit_mash_V(mash.data, Ulist, V, prior=prior)
    pi_s = get_estimated_pi(m.model, dimension = 'all')

    log_liks[i+1] <- get_loglik(m.model)+penalty(prior.v, pi_s)
    
    result = list(V = V, logliks = log_liks[1:(i+1)], mash.model = m.model)

    # Update delta
    delta.ll <- log_liks[i+1] - log_liks[i]
    if(delta.ll<=tol) break;
  }
  
  if(track_fit){
    result$trace = tracking
  }
  
  return(result)
}

E_V = function(mash.data, m.model){
  n = mashr:::n_effects(mash.data)
  Z = mash.data$Bhat/mash.data$Shat
  post.m.shat = m.model$result$PosteriorMean / mash.data$Shat
  post.sec.shat = plyr::laply(1:n, function(i) (t(m.model$result$PosteriorCov[,,i]/mash.data$Shat[i,])/mash.data$Shat[i,]) + tcrossprod(post.m.shat[i,])) # nx2x2 array
  temp1 = crossprod(Z)
  temp2 = crossprod(post.m.shat, Z) + crossprod(Z, post.m.shat)
  temp3 = unname(plyr::aaply(post.sec.shat, c(2,3), sum))

  (temp1 - temp2 + temp3)/n
}

fit_mash_V <- function(mash.data, Ulist, V, prior=c('nullbiased', 'uniform')){
  m.data = mashr::mash_set_data(Bhat=mash.data$Bhat, Shat=mash.data$Shat, V = V, alpha = mash.data$alpha)
  m.model = mashr::mash(m.data, Ulist, prior=prior, verbose = FALSE, outputlevel = 3)
  return(m.model)
}
```

## Data

```{r}
set.seed(1)
n = 4000; p = 2
Sigma = matrix(c(1,0.5,0.5,1),p,p)
U0 = matrix(0,2,2)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[2,2] = 1
U3 = matrix(1,2,2)
Utrue = list(U0=U0, U1=U1, U2=U2, U3=U3)
data = generate_data(n, p, Sigma, Utrue)
```

```{r}
m.data = mash_set_data(data$Bhat, data$Shat)
U.c = cov_canonical(m.data)

result.mV <- mixture.MV(m.data, U.c)
```

The estimated $V$ is 
```{r}
result.mV$V
```
The running time is `r result.mV$time` seconds.

```{r}
m.mV = result.mV$mash.model
null.ind = which(apply(data$B,1,sum) == 0)
```

The log likelihood is `r formatC(get_loglik(m.mV), digits=7)`. There are `r length(get_significant_results(m.mV))` significant samples, `r sum(get_significant_results(m.mV) %in% null.ind)` false positives. The RRMSE is `r RRMSE(data$B, data$Bhat, list(m.mV))`.
