---
title: "Estimate Null Correlation Problem"
author: "Yuxin Zou"
date: 2018-07-09
output: 
  workflowr::wflow_html:
    code_folding: hide
---

```{r library}
library(mashr)
library(knitr)
library(kableExtra)
source('../code/generateDataV.R')
source('../code/summary.R')
```

We illustrate the problem about estimating the correlation matrix in `mashr`. 

In my simple simulation, the current approach underestimates the null correlation. We want to find better positive definite estimator. We could try to estimate the pairwise correlation, ie. mle of $\sum_{l,k} \pi_{lk} N_{2}(0, V + w_{l}U_{k})$ for any pair of conditions.

## Problem

Simple simulation in $R^2$ to illustrate the problem:
$$
\hat{\beta}|\beta \sim N_{2}(\hat{\beta}; \beta, \left(\begin{matrix} 1 & 0.5 \\
                                          0.5 & 1 \end{matrix}\right))
$$

$$
\beta \sim \frac{1}{4}\delta_{0} + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 0 \\
                                          0 & 0 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 0 & 0 \\
                                          0 & 1 \end{matrix}\right)) + \frac{1}{4}N_{2}(0, \left(\begin{matrix} 1 & 1 \\
                                          1 & 1 \end{matrix}\right))
$$

$\Rightarrow$
$$
\hat{\beta} \sim \frac{1}{4}N_{2}(0, \left( \begin{matrix} 1 & 0.5 \\
0.5 & 1 \end{matrix} \right)) + \frac{1}{4}N_{2}(0, \left( \begin{matrix} 2 & 0.5 \\
0.5 & 1 \end{matrix} \right)) + \frac{1}{4}N_{2}(0, \left( \begin{matrix} 1 & 0.5 \\
0.5 & 2 \end{matrix} \right)) + \frac{1}{4}N_{2}(0, \left( \begin{matrix} 2 & 1.5 \\
1.5 & 2 \end{matrix} \right))
$$

n = 4000

```{r}
set.seed(1)
n = 4000; p = 2
Sigma = matrix(c(1,0.5,0.5,1),p,p)
U0 = matrix(0,2,2)
U1 = U0; U1[1,1] = 1
U2 = U0; U2[2,2] = 1
U3 = matrix(1,2,2)
Utrue = list(U0=U0, U1=U1, U2=U2, U3=U3)
data = generate_data(n, p, Sigma, Utrue)
```

Let's check the result of `mash` under different correlation matrix:

1. Identity
$$
V.I = I_{2}
$$

```{r}
m.data = mash_set_data(data$Bhat, data$Shat)
U.c = cov_canonical(m.data)
m.I = mash(m.data, U.c, verbose= FALSE)
```

2. The current approach: truncated empirical correlation, $V.trun$

The correlation of samples with maximum |z| scores less than 2.

```{r}
Vhat = estimate_null_correlation(m.data, apply_lower_bound = FALSE)
Vhat
```

It underestimates the correlation.

```{r}
# Use underestimate cor
m.data.V = mash_set_data(data$Bhat, data$Shat, V=Vhat)
m.V = mash(m.data.V, U.c, verbose = FALSE)
```

3. Overestimate correlation
$$
V.o = \left( \begin{matrix} 1 & 0.65 \\ 0.65 & 1\end{matrix}  \right)
$$
```{r}
# If we overestimate cor
V.o = matrix(c(1,0.65,0.65,1),2,2)
m.data.Vo = mash_set_data(data$Bhat, data$Shat, V=V.o)
m.Vo = mash(m.data.Vo, U.c, verbose=FALSE)
```

4. mash.1by1

We run ash for each condition, and estimate correlation matrix based on the non-significant genes. The estimated cor is closer to the truth.
```{r}
m.1by1 = mash_1by1(m.data)
strong = get_significant_results(m.1by1)
V.mash = cor(data$Bhat[-strong,])
V.mash
```

```{r}
m.data.1by1 = mash_set_data(data$Bhat, data$Shat, V=V.mash)
m.V1by1 = mash(m.data.1by1, U.c, verbose = FALSE)
```

5. True correlation
```{r}
# With correct cor
m.data.correct = mash_set_data(data$Bhat, data$Shat, V=Sigma)
m.correct = mash(m.data.correct, U.c, verbose = FALSE)
```

The results are summarized in table:
```{r}
null.ind = which(apply(data$B,1,sum) == 0)
V.trun = c(get_loglik(m.V), length(get_significant_results(m.V)), sum(get_significant_results(m.V) %in% null.ind))
V.I = c(get_loglik(m.I), length(get_significant_results(m.I)), sum(get_significant_results(m.I) %in% null.ind))
V.over = c(get_loglik(m.Vo), length(get_significant_results(m.Vo)), sum(get_significant_results(m.Vo) %in% null.ind))
V.1by1 = c(get_loglik(m.V1by1), length(get_significant_results(m.V1by1)), sum(get_significant_results(m.V1by1) %in% null.ind))
V.correct = c(get_loglik(m.correct), length(get_significant_results(m.correct)), sum(get_significant_results(m.correct) %in% null.ind))
temp = cbind(V.I, V.trun, V.1by1, V.correct, V.over)
colnames(temp) = c('Identity','truncate', 'm.1by1', 'true', 'overestimate')
row.names(temp) = c('log likelihood', '# significance', '# False positive')
temp %>% kable() %>% kable_styling()
```
<!--
The estimated `pi` is
```{r fig.align = "center"}
par(mfrow=c(2,3))
barplot(get_estimated_pi(m.I), las=2, cex.names = 0.7, main='Identity', ylim=c(0,0.8))
barplot(get_estimated_pi(m.V), las=2, cex.names = 0.7, main='Truncate', ylim=c(0,0.8))
barplot(get_estimated_pi(m.V1by1), las=2, cex.names = 0.7, main='m.1by1', ylim=c(0,0.8))
barplot(get_estimated_pi(m.correct), las=2, cex.names = 0.7, main='True', ylim=c(0,0.8))
barplot(get_estimated_pi(m.Vo), las=2, cex.names = 0.7, main='OverEst', ylim=c(0,0.8))
```
-->

The ROC curve:
```{r}
m.I.seq = ROC.table(data$B, m.I)
m.V.seq = ROC.table(data$B, m.V)
m.Vo.seq = ROC.table(data$B, m.Vo)
m.V1by1.seq = ROC.table(data$B, m.V1by1)
m.correct.seq = ROC.table(data$B, m.correct)
```

```{r echo=FALSE, fig.align = "center"}
{plot(m.correct.seq[,'FPR'], m.correct.seq[,'TPR'],type='l',xlab = 'FPR', ylab='TPR', main='True Positive vs False Positive', cex=1.5, lwd = 1.5)
lines(m.I.seq[,'FPR'], m.I.seq[,'TPR'], col='red', lwd = 1.5)
lines(m.V.seq[,'FPR'], m.V.seq[,'TPR'], col='green', lwd = 1.5)
lines(m.Vo.seq[,'FPR'], m.Vo.seq[,'TPR'], col='blue', lwd = 1.5)
lines(m.V1by1.seq[,'FPR'], m.V1by1.seq[,'TPR'], col='cyan', lwd = 1.5)
legend('bottomright', c('True','Identity','Trunc', 'OverEst', 'm.1by1'),col=c('black','red','green', 'blue', 'cyan'),lty=c(1,1,1,1,1), lwd=c(1.5,1.5,1.5,1.5,1.5))}
```

Comparing accuracy
```{r}
rrmse = rbind(RRMSE(data$B, data$Bhat, list(m.I = m.I, m.V = m.V, m.1by1 = m.V1by1, m.true = m.correct, m.over = m.Vo)))
colnames(rrmse) = c('Identity','V.trun','V.1by1','V.true','V.over')
row.names(rrmse) = 'RRMSE'
rrmse %>% kable() %>% kable_styling()
```
```{r}
barplot(rrmse, ylim=c(0,(1+max(rrmse))/2), las=2, cex.names = 0.7, main='RRMSE')
```

## Solution: MLE

### K=1

Suppose a simple extreme case
$$
\left(\begin{matrix} \hat{x} \\ \hat{y} \end{matrix} \right)| \left(\begin{matrix} x \\ y \end{matrix} \right) \sim N_{2}(\left(\begin{matrix} \hat{x} \\ \hat{y} \end{matrix} \right); \left(\begin{matrix} x \\ y \end{matrix} \right), \left( \begin{matrix} 1 & \rho \\ \rho & 1 \end{matrix}\right))
$$
$$
\left(\begin{matrix} x \\ y \end{matrix} \right) \sim \delta_{0}
$$
$\Rightarrow$
$$
\left(\begin{matrix} \hat{x} \\ \hat{y} \end{matrix} \right) \sim N_{2}(\left(\begin{matrix} \hat{x} \\ \hat{y} \end{matrix} \right); \left(\begin{matrix} 0 \\ 0 \end{matrix} \right), \left( \begin{matrix} 1 & \rho \\ \rho & 1 \end{matrix}\right))
$$

$$
f(\hat{x},\hat{y}) = \prod_{i=1}^{n} \frac{1}{2\pi\sqrt{1-\rho^2}} \exp \{-\frac{1}{2(1-\rho^2)}\left[ \hat{x}_{i}^2 + \hat{y}_{i}^2 - 2\rho \hat{x}_{i}\hat{y}_{i}\right]  \}
$$
The MLE of $\rho$:
$$
\begin{align*}
l(\rho) &= -\frac{n}{2}\log(1-\rho^2) - \frac{1}{2(1-\rho^2)}\left( \sum_{i=1}^{n} x_{i}^2 + y_{i}^2 - 2\rho x_{i}y_{i} \right) \\
l(\rho)' &= \frac{n\rho}{1-\rho^2} - \frac{\rho}{(1-\rho^2)^2} \sum_{i=1}^{n} (x_{i}^2 + y_{i}^2) + \frac{\rho^2 + 1}{(1-\rho^2)^2} \sum_{i=1}^{n} x_{i}y_{i} = 0 \\
&= \rho^{3} - \rho^{2}\frac{1}{n}\sum_{i=1}^{n} x_{i}y_{i} - \left( 1- \frac{1}{n} \sum_{i=1}^{n} x_{i}^{2} + y_{i}^{2} \right) \rho - \frac{1}{n}\sum_{i=1}^{n} x_{i}y_{i} = 0 \\
l(\rho)'' &= \frac{n(\rho^2+1)}{(1-\rho^2)^2} - \frac{1}{2}\left( \frac{8\rho^2}{(1-\rho^2)^{3}} + \frac{2}{(1-\rho^2)^2} \right)\sum_{i=1}^{n}(x_{i}^2 + y_{i}^2) + \{ \left( \frac{8\rho^2}{(1-\rho^2)^{3}} + \frac{2}{(1-\rho^2)^2} \right)\rho + \frac{4\rho}{(1-\rho^2)^2} \}\sum_{i=1}^{n}x_{i}y_{i}
\end{align*}
$$

**The log likelihood is not a concave function in general.** The score function has either 1 or 3 real solutions. 

Kendall and Stuart (1979) noted that at least one of the roots is real and lies in
the interval [−1, 1]. However, it is possible that all three roots are real and in the admissible interval, in which case the likelihood can be evaluated at each root to determine the true maximum likelihood estimate. 

I simulate the data with $\rho=0.6$ and plot the loglikelihood function:

```{r echo=FALSE}
n = 1000; rho = 0.6
Sigma = matrix(c(1,rho,rho,1),2,2)
Utrue = list(U0=matrix(0,2,2))
data = generate_data(n,2,Sigma, Utrue)

rho.seq = seq(-0.99,0.99,by=0.01)
loglik = numeric(199)
for(i in 1:199){
  V = matrix(c(1,rho.seq[i], rho.seq[i], 1), 2, 2)
  loglik[i] = sum(mvtnorm::dmvnorm(x=data$Bhat, sigma=V, log=TRUE))
}
plot(rho.seq, loglik, type='l', ylab = 'log likelihood')
```

$l(\rho)'$ has one real solution
```{r}
polyroot(c(- sum(data$Bhat[,1]*data$Bhat[,2]),  - (n - sum(data$Bhat[,1]^2 + data$Bhat[,2]^2)), - sum(data$Bhat[,1]*data$Bhat[,2]), n))
```

### In general

The general derivation is in [estimate correlation mle](EstimateCorOptim.html)

