---
title: "Presidential address-Flash"
author: "Yuxin Zou"
date: 2018-2-21
output: 
  html_document:
    code_folding: hide
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r, echo=FALSE}
# TEMPORARY.
knitr::opts_chunk$set(eval = TRUE)
```

```{r}
library(flashr); library(R.matlab)
library(denoiseR)
```

Presidential Address data:

The data contains word counts from the inaugural addresses of 13 US presidents (1940–2009). There are 13 rows (presidents) and 836 columns (words), and each entry indicates the number of times a particular word was used in a particular address. Since both row and column means vary greatly here we pre-processed the data by centering and scaling both rows and columns, using the biScale function from softImpute.

```{r}
data("Presidents")
data = readMat('../data/PresiAdd.mat')
data = data$Yscaled
words = row.names(Presidents)
names = colnames(Presidents)
row.names(data) = names
colnames(data) = words
```

Flash:
```{r}
flash.data = flash_set_data(data)
fmodel = flash(flash.data, greedy = TRUE, backfit = TRUE)

saveRDS(fmodel, '../output/PresiAddVarCol.rds')
```

```{r, echo=FALSE}
fmodel = readRDS('../output/PresiAddVarCol.rds')
```

```{r}
Factors = flash_get_ldf(fmodel)$f
row.names(Factors) = words
pve.order = order(flash_get_pve(fmodel), decreasing = TRUE)
```
```{r}
par(mar=c(1,1,1,1))
par(mfrow=c(3,2))
for(i in pve.order){
  barplot(Factors[,i], main=paste0('Factor ',i, ' pve= ', round(flash_get_pve(fmodel)[i],3)), las=2, names='')
}
par(mfrow=c(1,1))
```

Flash again on the loading matrix
```{r}
flash.loading = flash_set_data(fmodel$EL[,1:5])
flmodel = flash(flash.loading, greedy = TRUE, backfit = TRUE)
```
```{r}
Factors_loading = flash_get_ldf(flmodel)$f
```
```{r}
barplot(as.numeric(Factors_loading), main=paste0('Factor 1 pve= ', round(flash_get_pve(flmodel),3)), las=2, cex.names = 0.4, names='')
```

The distributions for the first four columns of the loading matrix are consistent with a normal distribution. So they are captured by the error term of the second flash run.

It is hard to interpret each factor, none of them is sparse. For example, for factor 1 in the first flash run,
```{r}
words[Factors[,1] > 0]
```

The first factor separate the words above from others. But it is hard to summarize the feature of these words.

The fifth factor separate the words below from others.
```{r}
words[Factors[,5] > 0.1]
```


# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
