---
title: "Presidential address-Flash"
author: "Yuxin Zou"
date: 2018-2-21
output: 
  html_document:
    code_folding: hide
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r, echo=FALSE}
# TEMPORARY.
knitr::opts_chunk$set(eval = TRUE)
```

```{r}
library(flashr); library(R.matlab)
library(denoiseR)
```

Presidential Address data:

The data contains word counts from the inaugural addresses of 13 US presidents (1940–2009). There are 13 rows (presidents) and 836 columns (words), and each entry indicates the number of times a particular word was used in a particular address. Since both row and column means vary greatly here we pre-processed the data by centering and scaling both rows and columns, using the biScale function from softImpute.

```{r}
data("Presidents")
data = readMat('../data/PresiAdd.mat')
data = data$Yscaled
words = row.names(Presidents)
names = colnames(Presidents)
row.names(data) = names
colnames(data) = words
```

Flash:
```{r}
flash.data = flash_set_data(data)
fmodel = flash(flash.data, greedy = TRUE, backfit = TRUE)

saveRDS(fmodel, '../output/PresiAddVarCol.rds')
```

```{r, echo=FALSE}
fmodel = readRDS('../output/PresiAddVarCol.rds')
```

```{r}
Factors = flash_get_ldf(fmodel)$f
row.names(Factors) = words
pve.order = order(flash_get_pve(fmodel), decreasing = TRUE)
```
```{r}
par(mar=c(1,1,1,1))
par(mfrow=c(3,2))
for(i in pve.order){
  barplot(Factors[,i], main=paste0('Factor ',i, ' pve= ', round(flash_get_pve(fmodel)[i],3)), las=2, names='')
}
par(mfrow=c(1,1))
```

Flash again on the loading matrix
```{r}
flash.loading = flash_set_data(fmodel$EL[,1:5])
flmodel = flash(flash.loading, greedy = TRUE, backfit = TRUE)
```
```{r}
Factors_loading = flash_get_ldf(flmodel)$f
```
```{r}
barplot(as.numeric(Factors_loading), main=paste0('Factor 1 pve= ', round(flash_get_pve(flmodel),3)), las=2, cex.names = 0.4, names='')
```

The distributions for the first four columns of the loading matrix are consistent with a normal distribution. So they are captured by the error term of the second flash run.

We check the strongest words in each factor.

The first factor separates the words about Iraq from those very general political words.
```{r}
# Factor 1
words[order(Factors[,1], decreasing = TRUE)[1:50]]
words[order(Factors[,1])[1:50]]
```

The second factor separates the words about people's life, like 'child', 'job', 'school', 'parent', 'Medicare', from political words.

```{r}
# Factor 2
words[order(Factors[,2], decreasing = TRUE)[1:50]]
words[order(Factors[,2])[1:50]]
```

The third factor separate the words about government and development of technology. These words are used in Carter's address.

```{r}
# Factor 3
words[order(Factors[,3], decreasing = TRUE)[1:50]]
words[order(Factors[,3])[1:50]]
```

The fourth factor separate the words about Vietnam war.
```{r}
# Factor 4
words[order(Factors[,4], decreasing = TRUE)[1:50]]
words[order(Factors[,4])[1:50]]
```

The fifth factor separates words about Vietnam war from words like 'peacetime', 'reform'.  

```{r}
# Factor 5
words[order(Factors[,5], decreasing = TRUE)[1:50]]
words[order(Factors[,5])[1:50]]
```

Checking the loading matrix
```{r}
loading = fmodel$EL[,1:5]
row.names(loading) = names
loading
```

Kennedy, Johnson and Nixon used the words in Factor 5 a lot. This is reasonable since the Vietnam war was happened during their presidency term. When Nixon took office, people want peace because of the long times of the war. So he used words about 'peace' in his address a lot.


# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
