---
title: "Brain Tumor-Flash"
author: "Yuxin Zou"
date: 2018-2-20
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r, echo=FALSE}
# TEMPORARY.
knitr::opts_chunk$set(eval = TRUE)
```

```{r echo=FALSE}
library(flashr); library(R.matlab)
```

Brain Tumor data:

The data consist of micro array gene expression measurements on 43 brain tumors of four types, together with 356 additional variables related to the expression data. We pre-processed the data by centering the data by column.
```{r}
data = readMat('../data/BrainTumor.mat')
data = data$Y
```

Flash:
```{r, eval=FALSE}
flash.data = flash_set_data(data)
f = flash(flash.data, greedy = TRUE, backfit = TRUE)

saveRDS(f, '../output/BrainTumorVarCol.rds')
```

```{r, echo=FALSE}
fmodel = readRDS('../output/BrainTumorVarCol.rds')
```

```{r, echo=FALSE}
Factors = flash_get_ldf(fmodel)$f
pve.order = order(flash_get_pve(fmodel), decreasing = TRUE)
```
```{r, echo=FALSE}
par(mar=c(1,1,1,1))
par(mfrow=c(3,2))

for(i in pve.order){
  barplot(Factors[,i], main=paste0('Factor ',i, ' pve= ', round(flash_get_pve(fmodel)[i],3)), las=2, cex.names = 0.4, names="")
}
```

Flash again on the loading matrix
```{r, eval=FALSE}
flash.loading = flash_set_data(fmodel$EL[,1:9])
flmodel = flash(flash.loading, greedy = TRUE, backfit = TRUE)
```

The flash prefers the rank 0 model. There is no hidden structure in the loading matrix.

# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
