---
title: "MovieLens 100K-Flash"
author: "Yuxin Zou"
date: 2018-2-21
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r, echo=FALSE}
# TEMPORARY.
knitr::opts_chunk$set(eval = TRUE)
```

```{r echo=FALSE}
library(flashr)
library(methods)
library(R.matlab)
library(Matrix)
```

MovieLens 100K data:

The data contains about 100K user-movie ratings (integers from 1 to 5) from approximately 1000 users across a total of 1700 movies. Most users do not rate most movies, so the matrix is sparsely observed (94% missing). We centered and scaled the ratings for each user before applying methods to these data.
```{r}
ml100K_data = readRDS("../data/scale_data.rds")
MLMatrix <- sparseMatrix(i = ml100K_data[,1],
                         j = ml100K_data[,2],
                         x = ml100K_data[,3],dims = c(943,1682))
# turn this sparse matrix into matrix in r
Y = as.matrix(MLMatrix)
Y[which(Y == 0)] = NA
```

Flash:
```{r, eval=FALSE}
flash_data = flash_set_data(Y)
fmodel = flash(flash_data, greedy = TRUE, backfit = TRUE)

saveRDS(fmodel, '../output/MovieLensVarCol.rds')
```

```{r, echo=FALSE}
fmodel = readRDS('../output/MovieLensVarCol.rds')
```

```{r, echo=FALSE}
Factors = flash_get_ldf(fmodel)$f
```
```{r, echo=FALSE}
par(mar=c(1,1,1,1))
par(mfrow=c(3,2))
for(i in 1:5){
  barplot(Factors[,i], main=paste0('Factor ',i, ' pve= ', round(flash_get_pve(fmodel)[i],3)), las=2, cex.names = 0.4, names="")
}
```

Flash again on the loading matrix
```{r, eval=FALSE}
flash_loading = flash_set_data(fmodel$EL[,1:5])
flmodel = flash(flash_loading, ebnm_fn = ebnm_ash, backfit = TRUE, greedy = TRUE)
```

The flash prefers the rank 0 model. There is no hidden structure in the loading matrix.

# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
