---
title: "UKBioBank MASH"
author: "Yuxin Zou"
date: 2018-6-27
output: 
  html_document:
    code_folding: hide
---
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r echo=FALSE}
library(flashr); library(mashr);library(mclust);library(plyr)
library(lattice);library(ggplot2);library(colorRamps)
```

UKBioBank Strong data
```{r}
data = readRDS('../data/UKBioBank/StrongData.rds')
```

Estimate se based on *p values*
```{r}
# Adjust p value == 0
data$p[data$p == 0] = 1e-323
```

Fit EZ model directly to the Z scores, with standard errors of the non-missing Z scores set to 1 and the missing ones set to 10^6.
```{r}
mash.data = mash_set_data(Bhat = data$beta, pval = data$p, alpha = 1)
Zhat = mash.data$Bhat; Shat = mash.data$Shat
missing = is.na(Zhat)
Shat[missing] = 10^6
Zhat[missing] = 0
data.EZ = mash_set_data(Zhat,Shat)
```

# Estimate Covariance

```{r}
# column center
Z.center = apply(Zhat, 2, function(x) x - mean(x))
```

## Flash on centered Z

$$
Z = LF' + E
$$
Z is an $n \times p$ observed centered data, F is a $p\times k$ matrix of factors, L is an $n \times k$ matrix of loadings.

Results from [Flash_UKBio](Flash_UKBio.html)
```{r}
fmodel = readRDS('../output/Flash_UKBio_strong.rds')
```

Suppose the rows of L come from a mixture of multivariate normals, with covariances $\Sigma_1,\dots,\Sigma_M$ (each one a K by K matrix).
$$
l_{i \cdot} \sim \sum_{j=1}^{M} N(\mu_{j}, \Sigma_{j})
$$
Then the rows of $LF'$ come from a mixture of multivariate normals 
$$
Fl_{i\cdot} \sim \sum_{j=1}^{M} N(F\mu_{j}, F\Sigma_{j}F')
$$
We estimate the covariance matrix as $F(\Sigma_{j}+\mu_{j}\mu_{j}')F′$.

Cluster loadings:
```{r}
loading = fmodel$EL[,1:18]
colnames(loading) = paste0('Factor',seq(1,18))
mod = Mclust(loading)
summary(mod$BIC)
```
```{r}
U_list = alply(mod$parameters$variance$sigma,3)
mu_list = alply(mod$parameters$mean,2)
ll = list()
for (i in 1:length(U_list)){
  ll[[i]] = U_list[[i]] + mu_list[[i]] %*% t(mu_list[[i]])
}

Factors = fmodel$EF[,1:18]
U.loading = lapply(ll, function(U){Factors %*% (U %*% t(Factors))})
names(U.loading) = paste0('Load', "_", (1:length(U.loading)))

# rank 1
Flash_res = flash_get_lf(fmodel)
U.Flash = c(mashr::cov_from_factors(t(as.matrix(Factors)), "Flash"), 
            list("tFlash" = t(Flash_res) %*% Flash_res / nrow(Z.center)))
```

## PCA
```{r}
U.pca = cov_pca(data.EZ, 5)
```

## Extreme Deconvolution

```{r, eval=FALSE}
U.dd = c(U.pca, U.loading, U.Flash, list('XX' = t(data.EZ$Bhat) %*% data.EZ$Bhat / nrow(data.EZ$Bhat)))

U.ed = cov_ed(data.EZ, U.dd)
saveRDS(U.ed, '../output/CovED_UKBio_strong_Z.rds')
```

```{r}
U.ed = readRDS('../output/CovED_UKBio_strong_Z.rds')
```

## Canonical
```{r}
U.c = cov_canonical(data.EZ)
```

# Mash model

Read random data
```{r}
data.rand = readRDS('../data/UKBioBank/RandomData.rds')
# Estimate se based on p values
# Adjust p value == 0
data.rand$p[data.rand$p == 0] = 1e-323
mash.data.rand = mash_set_data(Bhat = data.rand$beta, pval = data.rand$p, alpha = 1)
Zhat = mash.data.rand$Bhat; Shat = mash.data.rand$Shat
missing = is.na(Zhat)
Shat[missing] = 10^6
Zhat[missing] = 0
data.rand.EZ = mash_set_data(Zhat,Shat)
Vhat = estimate_null_correlation(data.rand.EZ)
```

```{r, eval=FALSE}
data.rand.EZ.V = mash_set_data(data.rand.EZ$Bhat, data.rand.EZ$Shat, V = Vhat)
mash.model = mash(data.rand.EZ.V, c(U.c, U.ed), outputlevel = 1)
saveRDS(mash.model, '../output/UKBio_mash_model.rds')
```

```{r}
mash.model = readRDS('../output/UKBio_mash_model.rds')
```

The log-likelihood of fit is
```{r} 
get_loglik(mash.model)
```

Here is a plot of weights learned:

```{r}
options(repr.plot.width=12, repr.plot.height=4)
barplot(get_estimated_pi(mash.model), las = 2, cex.names = 0.7)
```

The founded correlations:

From the empirical correlation matrix, the metabolic disorder diseases are correlated (BMI, Blood Pressure, Cardiovascular and Type 2 Diabetes), which agrees the Factor 2 in the flash result. The eosinophil cell count is positively correlated with white cell count, since eosinophil cell is a variety of white blood cells. Eosinophil cell cpunt is also positively correlated with respiratory disease. Having respiratory disease decreases FEV1FVC ratio. Years of education is negatively correlated with smoking status. There must be some strong environmental factors that are affecting these phenotypes together. The associations from the mash result here may not the pure genetic associations. Moreover, from the `load_4` covariance matrix, neuroticism and smoking are highly correlated here (0.71). Neurotic people may smoke more and finish their school life earlier.



`ED_XX`:
```{r ED_XX, fig.height = 15, fig.width = 15, fig.align = "center"}
x           <- cov2cor(mash.model$fitted_g$Ulist[["ED_XX"]])
colnames(x) <- colnames(get_lfsr(mash.model))
rownames(x) <- colnames(x)
corrplot::corrplot(x, method='color', cl.lim=c(-1,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF", "#E0F3F8","#91BFDB","#4575B4")))(128))
```


The top eigenvalues:
```{r}
svd.out = svd(mash.model$fitted_g$Ulist[["ED_XX"]])
v = svd.out$v
colnames(v) = colnames(get_lfsr(mash.model))
rownames(v) = colnames(v)
options(repr.plot.width=10, repr.plot.height=5)
for (j in 1:6)
  barplot(v[,j]/v[,j][which.max(abs(v[,j]))], cex.names = 0.6,
          las = 2, main = paste0("EigenVector ", j, " for empirical covariance matrix"))
```

`ED_load_4`
```{r ED_load_4, fig.height = 15, fig.width = 15, fig.align = "center"}
x           <- cov2cor(mash.model$fitted_g$Ulist[["ED_Load_4"]])
colnames(x) <- colnames(get_lfsr(mash.model))
rownames(x) <- colnames(x)
corrplot::corrplot(x, method='color', cl.lim=c(-1,1), type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, col=colorRampPalette(rev(c("#D73027","#FC8D59","#FEE090","#FFFFBF", "#E0F3F8","#91BFDB","#4575B4")))(128))
```

The top eigenvalues:
```{r}
svd.out = svd(mash.model$fitted_g$Ulist[["ED_Load_4"]])
v = svd.out$v
colnames(v) = colnames(get_lfsr(mash.model))
rownames(v) = colnames(v)
options(repr.plot.width=10, repr.plot.height=5)
for (j in 1:3)
  barplot(v[,j]/v[,j][which.max(abs(v[,j]))], cex.names = 0.6,
          las = 2, main = paste0("EigenVector ", j, " for Load 4 covariance matrix"))
```

# Posterior

```{r}
data.strong = mash_set_data(data.EZ$Bhat, data.EZ$Shat, V = Vhat)
mash.model$result = mash_compute_posterior_matrices(mash.model, data.strong)
```

There are `r length(get_significant_results(mash.model))` significant snps.

Pairwise sharing (Shared by magnitude when sign is ignored):

```{r sharing by magnitude, fig.align='center', fig.width=15, fig.height=15}
x           <- get_pairwise_sharing(mash.model, FUN = abs)
colnames(x) <- colnames(get_lfsr(mash.model))
rownames(x) <- colnames(x)
clrs=colorRampPalette(rev(c('darkred', 'red','orange','yellow','cadetblue1', 'cyan', 'dodgerblue4', 'blue','darkorchid1','lightgreen','green', 'forestgreen','darkolivegreen')))(200)

corrplot::corrplot(x, method='color', type='upper', addCoef.col = "black", tl.col="black", tl.srt=45, diag = FALSE, col=clrs, cl.lim = c(0,1))
```

Pairwise sharing (Shared by magnitude and sign):

```{r sharing by magnitude and sign, fig.align='center', fig.width=15, fig.height=15}
x           <- get_pairwise_sharing(mash.model)
colnames(x) <- colnames(get_lfsr(mash.model))
rownames(x) <- colnames(x)
clrs=colorRampPalette(rev(c('darkred', 'red','orange','yellow','cadetblue1', 'cyan', 'dodgerblue4', 'blue','darkorchid1','lightgreen','green', 'forestgreen','darkolivegreen')))(200)
corrplot::corrplot(x, method='color', type='upper', addCoef.col = "black", tl.col="black", diag=FALSE,tl.srt=45, col=clrs, cl.lim = c(0,1))
```

# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```

