---
title: "Mash Missing Whole Row"
author: "Yuxin Zou"
date: 2018-2-28
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r, echo=FALSE}
# TEMPORARY.
knitr::opts_chunk$set(eval = TRUE)
```

# R=5

```{r}
library(mashr)
set.seed(2018)
data = simple_sims(500, err_sd = 0.1)

data.miss = data
missing.row = sample(1:nrow(data$Bhat), nrow(data$Bhat)/4)
missing.ind = matrix(0, nrow = nrow(data$B), ncol=ncol(data$B))
missing.ind[missing.row,] = 1

missing = which(missing.ind == 1)
data.miss$Bhat[missing] = NA
data.miss$Shat[missing] = NA
missing1 = is.na(data.miss$Bhat[,1])
# Set the missing value with large standard deviation
data.large.dev = data.miss
data.large.dev$Bhat[is.na(data.miss$Bhat)] = 0
data.large.dev$Shat[is.na(data.miss$Shat)] = 1000
```

## EE model

```{r, echo=FALSE}
# with missing values
mash.data.missing = mash_set_data(Bhat=data.large.dev$Bhat, Shat=data.large.dev$Shat)
U.c = cov_canonical(mash.data.missing)

mash.model.missing = mash(mash.data.missing, U.c, verbose = FALSE)

# without missing values
data.miss.na = data.miss
data.miss.na$Bhat = na.omit(data.miss$Bhat)
data.miss.na$Shat = na.omit(data.miss$Shat)

mash.data.missing.na = mash_set_data(Bhat=data.miss.na$Bhat, Shat= data.miss.na$Shat)
U.c = cov_canonical(mash.data.missing.na)

mash.model.missing.na = mash(mash.data.missing.na, U.c, verbose=FALSE)
```

```{r, echo=FALSE}
par(mfrow=c(1,2))
barplot(get_estimated_pi(mash.model.missing), las=2, cex.names = 0.7, main='with missing values')
barplot(get_estimated_pi(mash.model.missing.na), las=2, cex.names = 0.7, main = 'without missing values')
par(mfrow=c(1,1))
```

## EZ model

```{r, echo=FALSE}
# With missing values
mash.data.missing.1 = mash_set_data(Bhat=data.large.dev$Bhat, Shat=data.large.dev$Shat, alpha = 1)
U.c = cov_canonical(mash.data.missing.1)

mash.model.missing.1 = mash(mash.data.missing.1, U.c, verbose = FALSE)

# Without missing values
mash.data.missing.na.1 = mash_set_data(Bhat=data.miss.na$Bhat, Shat= data.miss.na$Shat, alpha = 1)
U.c = cov_canonical(mash.data.missing.na.1)

mash.model.missing.na.1 = mash(mash.data.missing.na.1, U.c, verbose=FALSE)
```


```{r, echo=FALSE}
par(mfrow=c(1,2))
barplot(get_estimated_pi(mash.model.missing.1), las=2, cex.names = 0.7, main='with missing values')
barplot(get_estimated_pi(mash.model.missing.na.1), las=2, cex.names = 0.7, main='without missing values')
par(mfrow=c(1,1))
```

# R=60
```{r, echo=FALSE}
set.seed(2018)
data = simple_sims(500, ncond = 60, err_sd = 0.1)

data.miss = data
missing.row = sample(1:nrow(data$Bhat), nrow(data$Bhat)/4)
missing.ind = matrix(0, nrow = nrow(data$B), ncol=ncol(data$B))
missing.ind[missing.row,] = 1

missing = which(missing.ind == 1)
data.miss$Bhat[missing] = NA
data.miss$Shat[missing] = NA
missing1 = is.na(data.miss$Bhat[,1])
# Set the missing value with large standard deviation
data.large.dev = data.miss
data.large.dev$Bhat[is.na(data.miss$Bhat)] = 0
data.large.dev$Shat[is.na(data.miss$Shat)] = 1000
```

## EE model

### With missing values
```{r, echo=FALSE}
mash.data.missing = mash_set_data(Bhat=data.large.dev$Bhat, Shat=data.large.dev$Shat)
U.c = cov_canonical(mash.data.missing)

mash.model.missing = mash(mash.data.missing, U.c, verbose = FALSE)
barplot(get_estimated_pi(mash.model.missing), las=2, cex.names = 0.7)
```
The weights learned from the data are correct.

### Without missing values
```{r, echo=FALSE}
data.miss.na = data.miss
data.miss.na$Bhat = na.omit(data.miss$Bhat)
data.miss.na$Shat = na.omit(data.miss$Shat)

mash.data.missing.na = mash_set_data(Bhat=data.miss.na$Bhat, Shat= data.miss.na$Shat)
U.c = cov_canonical(mash.data.missing.na)

mash.model.missing.na = mash(mash.data.missing.na, U.c, verbose=FALSE)
barplot(get_estimated_pi(mash.model.missing.na), las=2, cex.names = 0.7)
```
The estimated weights are very weird and incorrect. It ignores the null matrix.

## EZ model

### With missing values
```{r, echo=FALSE}
mash.data.missing.1 = mash_set_data(Bhat=data.large.dev$Bhat, Shat=data.large.dev$Shat, alpha = 1)
U.c = cov_canonical(mash.data.missing.1)

mash.model.missing.1 = mash(mash.data.missing.1, U.c, verbose = FALSE)
barplot(get_estimated_pi(mash.model.missing.1), las=2, cex.names = 0.7)
```

### Without missing values
```{r}
mash.data.missing.na.1 = mash_set_data(Bhat=data.miss.na$Bhat, Shat= data.miss.na$Shat, alpha = 1)
U.c = cov_canonical(mash.data.missing.na.1)

mash.model.missing.na.1 = mash(mash.data.missing.na.1, U.c, verbose=FALSE)
barplot(get_estimated_pi(mash.model.missing.na.1), las=2, cex.names = 0.7)
```
The estimated weights are different.

# Subgroups

The weired covariance structure is caused by the high dimension of R.

If we separate conditions into several groups, the covariance structure will be correct.

We show this for the EZ model R=60.

```{r, echo=FALSE}
mash.data.missing.1.sub1 = mash_set_data(Bhat=data.large.dev$Bhat[,1:20], Shat=data.large.dev$Shat[,1:20], alpha = 1)
U.c = cov_canonical(mash.data.missing.1.sub1)
mash.model.missing.1.sub1 = mash(mash.data.missing.1.sub1, U.c, verbose = FALSE)

mash.data.missing.1.sub2 = mash_set_data(Bhat=data.large.dev$Bhat[,21:40], Shat=data.large.dev$Shat[,21:40], alpha = 1)
U.c = cov_canonical(mash.data.missing.1.sub2)
mash.model.missing.1.sub2 = mash(mash.data.missing.1.sub2, U.c, verbose = FALSE)

mash.data.missing.1.sub3 = mash_set_data(Bhat=data.large.dev$Bhat[,41:60], Shat=data.large.dev$Shat[,41:60], alpha = 1)
U.c = cov_canonical(mash.data.missing.1.sub3)
mash.model.missing.1.sub3 = mash(mash.data.missing.1.sub3, U.c, verbose = FALSE)

par(mfcol=c(3,1))
barplot(get_estimated_pi(mash.model.missing.1.sub1), las=2, cex.names = 0.7, main='condition 1:20')
barplot(get_estimated_pi(mash.model.missing.1.sub2), las=2, cex.names = 0.7, main='condition 21:40')
barplot(get_estimated_pi(mash.model.missing.1.sub3), las=2, cex.names = 0.7,
main='condition 41:60')
par(mfcol=c(1,1))
```
 
# Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
